### 2014-04-17 bayespy.tex
Initial commit

\documentclass[twoside,11pt]{article}

% Any additional packages needed should be included after jmlr2e.
% Note that jmlr2e.sty includes epsfig, amssymb, natbib and graphicx,
% and defines many common macros, such as 'proof' and 'example'.
%
% It also sets the bibliographystyle to plainnat; for more information on
% natbib citation styles, see the natbib documentation, a copy of which
% is archived at http://www.jmlr.org/format/natbib.pdf

\usepackage{jmlr2e}

% Definitions of handy macros can go here

\newcommand{\dataset}{{\cal D}}
\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}

% Heading arguments are {volume}{year}{pages}{submitted}{published}{author-full-names}

\jmlrheading{?}{2014}{?-?}{?/??}{?/??}{Jaakko Luttinen}

% Short headings should be running head and authors last names

\ShortHeadings{BayesPy: Variational Bayesian Inference in Python}{Luttinen}
\firstpageno{1}

\begin{document}

\title{BayesPy: Variational Bayesian Inference in Python}

\author{\name Jaakko Luttinen \email jaakko.luttinen@aalto.fi \\
       \addr Department of Information and Computer Science\\
       Aalto University, Finland}

\editor{?}

\maketitle

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
  This paper describes BayesPy.
\end{abstract}

\begin{keywords}
  Bayesian Networks
\end{keywords}

\section{Introduction}

Blaa blaa

% Acknowledgements should go at the end, before appendices and references

\acks{?}

\vskip 0.2in
\bibliography{bibliography.bib}

\end{document}

### 2014-08-07 bayespy.tex
Fix references
Start writing introduction

@@
 \usepackage{jmlr2e}
+\usepackage{doi}
 
 % Definitions of handy macros can go here
-
-\newcommand{\dataset}{{\cal D}}
-\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}
 
@@
 
-Blaa blaa
+Bayesian framework provides theoretically solid and consistent way to construct
+models and perform inference.  In practice, however, the inference is usually
+analytically intractable and is therefore based on approximation methods such as
+variational Bayes (VB) \cite{?}, expectation propagation (EP) \cite{?} and
+Markov chain Monte Carlo (MCMC) \cite{?}.  Deriving and implementing the
+formulas for the approximation method is often straightforward but tedious, time
+consuming and error prone.  Thus, automatization of this process ..
+
+BayesPy is a Python package providing tools for constructing Bayesian models and
+performing variational Bayesian inference easily and efficiently.  It is based
+on variational message passing (VMP) framework which defines a simple and local
+message passing protocol \cite{?}.  This enables implementation of small general
+modules that can be used as building blocks for large complex models.  BayesPy
+offers a comprehensive collection of built-in blocks that can be used to build a
+wide range of models.  It is written for Python 3 and released under the GNU
+General Public License v3.0 (GPLv3).
+
+Several other projects have similar goals for making Bayesian inference easier
+and faster for the users.  VB inference is available in Bayes Blocks \cite{?},
+VIBES \cite{?} and Infer.NET \cite{?}.  Bayes Blocks is an open-source
+C++/Python package but it is limited to scalar Gaussian nodes, a few nonlinear
+functions and fully factorial posterior approximation.  VIBES is an old software
+package for Java, released under the revised BSD license, but it is no longer
+actively developed.  VIBES has been replaced by Infer.NET, which supports
+
+\section{Design and Features}
+
+- numpy, scipy, matplotlib, h5py
+
+- documentation
+
+- installation
+
+- repository
+
+- unit tests
+
+- parameter expansion
+
+- missing values
+
+- examples: PCA, LSSM, HMM, mixture models
+
+Blaa blaa \cite{Luttinen:2013}.
+
+\section{Example}
+
+- example
+
+\section{Conclusion and Future Work}
+
+- future work
+-- ep, gibbs
+-- ml, map
 
@@
 \vskip 0.2in
-\bibliography{bibliography.bib}
+\bibliography{bibliography/bibliography.bib}
 
 \end{document}
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-PDF-mode: t
+%%% TeX-master: t
+%%% End:

### 2014-08-13 bayespy.tex
First version of the paper.

--- 2014-04-17 bayespy.tex
@@
 \usepackage{jmlr2e}
+\usepackage{doi}
+
+\usepackage{subcaption}
+\usepackage{url}
+\usepackage{color}
+\usepackage{listings}
+\usepackage{amsmath}
+\usepackage{tikz}
+\usetikzlibrary{bayesnet}
 
 % Definitions of handy macros can go here
-
-\newcommand{\dataset}{{\cal D}}
-\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}
 
@@
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-  This paper describes BayesPy.
+  BayesPy is an open-source Python software package for performing variational
+  Bayesian inference.  It supports conjugate exponential family models by
+  utilizing the variational message passing framework.  By removing the tedious
+  task of implementing the variational Bayesian update equations, the user can
+  construct models faster and in a less error-prone way.  Simple syntax,
+  flexible model construction and efficient inference make BayesPy suitable for
+  both average and expert Bayesian users.
 \end{abstract}
@@
 \begin{keywords}
-  Bayesian Networks
+  variational Bayes
 \end{keywords}
@@
 
-Blaa blaa
+Bayesian framework provides theoretically solid and consistent way to construct
+models and perform inference.  In practice, however, the inference is usually
+analytically intractable and is therefore based on approximation methods such as
+variational Bayes (VB), expectation propagation (EP) and Markov chain Monte
+Carlo (MCMC) \citep{Bishop:2006}.  Deriving and implementing the formulas for an
+approximation method is often straightforward but tedious, time consuming and
+error prone.
+
+BayesPy is a Python package providing tools for constructing Bayesian models and
+performing variational Bayesian inference easily and efficiently.  It is based
+on variational message passing (VMP) framework which defines a simple and local
+message passing protocol \citep{Winn:2005}.  This enables implementation of small
+general modules that can be used as building blocks for large complex models.
+BayesPy offers a comprehensive collection of built-in blocks that can be used to
+build a wide range of models and a simple interface for implementing custom
+blocks.  The package is written for Python 3 and released under the GNU General
+Public License v3.0 (GPLv3).
+
+Several other projects have similar goals for making Bayesian inference easier
+and faster to apply.  VB inference is available in Bayes Blocks
+\citep{Raiko:2007}, VIBES \citep{Bishop:2002} and Infer.NET \citep{Infer.NET}.
+Bayes Blocks is an open-source C++/Python package but it is limited to scalar
+Gaussian nodes, a few deterministic functions and fully factorial posterior
+approximation, thus making it very limited.  VIBES is an old software package
+for Java, released under the revised BSD license, but it is no longer actively
+developed.  VIBES has been replaced by Infer.NET, which supports a wide range of
+posterior approximations.  However, Infer.NET is partly closed source and
+licensed for non-commercial use only.  Instead of VB inference, mainly MCMC is
+supported by other projects such as PyMC \citep{PyMC}, OpenBUGS \citep{OpenBUGS},
+Dimple \citep{Dimple} and Stan \citep{Stan}.  Thus, there is a need for an open
+source and maintained VB software package.
+
+\section{Design and Features}
+
+BayesPy can be used to construct a wide range of conjugate exponential family
+models.  The documentation provides detailed examples of how to construct, for
+instance, regression models, principal component analysis models, linear
+state-space models, mixture models and hidden Markov models.  In addition,
+BayesPy has been used in two publications about parameter expansion and
+time-varying dynamics for linear state-space models
+\citep{Luttinen:2013,Luttinen:2014}, which show that it is applicable to large
+complex models requiring efficient inference.
+
+Using BayesPy for Bayesian inference consists of four main steps: constructing
+the model, providing data, finding the posterior approximation and examining the
+results.  The user constructs the model from small modular blocks called nodes.
+Roughly, each node corresponds to a latent variable, a set of observations or a
+deterministic function.  Some of the nodes may be set as observed by providing
+data.  After that, inference engine is initialized and the message passing
+algorithm is run in order to obtain the posterior approximation.  The resulting
+posterior can be examined, for instance, by using a few built-in plotting
+functions or printing the parameters of the posterior distributions.
+
+Nodes are the key part of BayesPy as they are used to construct the model.
+There are two types of nodes in BayesPy: stochastic and deterministic.
+Stochastic nodes correspond to probability distributions and deterministic nodes
+correspond to functions.  Built-in stochastic nodes include all common
+exponential family distributions (e.g., Gaussian, gamma, Dirichlet and Poisson
+distributions), a general mixture distribution and a few complex nodes for
+dynamic variables (e.g., discrete and Gaussian Markov chains).  Built-in
+deterministic nodes include a gating node and a general sum-product node.  In
+case a model cannot be constructed using the built-in nodes, the documentation
+provides instructions for implementing new nodes.
+
+BayesPy is designed to be simple enough for average users but flexible and
+efficient enough for expert users.  One goal is to keep the syntax easy and
+intuitive to read and write by making it similar to the mathematical formulation
+of the model.  Other notable features of BayesPy include easy handling of
+missing values, parameter expansions \citep{Qi:2007} for speeding up inference
+and monitoring of the variables by plotting the posterior distributions during
+the message passing iteration.  For developers, the unit testing framework helps
+in finding bugs, making changes and implementing new features in a robust
+manner.
+
+BayesPy can be installed similarly to other Python packages.  It requires Python
+3 and a few popular packages: NumPy, SciPy, matplotlib and h5py.  The latest
+release can be installed from Python Package Index (PyPI) and detailed
+instructions can be found from the comprehensive online documentation
+\citep{bayespy.org}.  The latest development version is available at
+GitHub\footnote{\url{https://github.com/bayespy/bayespy}}, which is also the
+platform used for reporting bugs and making pull requests.
+
+%- numpy, scipy, matplotlib, h5py
+
+%- documentation
+
+%- installation
+
+%- repository
+
+%- clear syntax
+
+%- unit tests
+
+%- parameter expansion
+
+%- monitoring
+
+%- missing values
+
+%- examples: PCA, LSSM, HMM, mixture models
+
+%- used in two publications, \citep{Luttinen:2013, Luttinen:2014}
+
+\section{Example}
+
+\lstset{ %
+  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
+  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+  breaklines=true,                 % sets automatic line breaking
+  captionpos=b,                    % sets the caption-position to bottom
+  commentstyle=\color{green},    % comment style
+  deletekeywords={...},            % if you want to delete keywords from the given language
+  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
+  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
+%  frame=single,                    % adds a frame around the code
+  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+  keywordstyle=\color{blue},       % keyword style
+  morekeywords={*,...},            % if you want to add more keywords to the set
+  firstnumber=last,
+  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
+  numbersep=10pt,                   % how far the line-numbers are from the code
+  numberstyle=\tiny\color{black}, % the style that is used for the line-numbers
+  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
+  showstringspaces=false,          % underline spaces within strings only
+  showtabs=false,                  % show tabs within strings adding particular underscores
+  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
+  stringstyle=\color{orange},     % string literal style
+  tabsize=2,                       % sets default tabsize to 2 spaces
+%  title=\lstname,                   % show the filename of files included with \lstinputlisting; also try caption instead of title
+  belowcaptionskip=1\baselineskip,
+  breaklines=true,
+  frame=L,
+  xleftmargin=1.3\parindent,
+  showstringspaces=false,
+  basicstyle=\footnotesize\ttfamily,
+%  keywordstyle=\bfseries\color{green!40!black},
+%  commentstyle=\itshape\color{purple!40!black},
+%  identifierstyle=\color{blue},
+%  stringstyle=\color{orange},
+  language=Python
+}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node [obs] (yn) {$y_n$};
+    \node [latent, above left=of yn] (mu) {$\mu$};
+    \node [latent, above right=of yn] (tau) {$\tau$};
+
+    \edge {mu, tau} {yn};
+
+    \tikzstyle{plate caption} = [caption, node distance=0, inner sep=0pt,
+    below=5pt of #1.south]
+
+    \plate {} {(yn)} {$n=1,\ldots,10$};
+  \end{tikzpicture}
+  \caption{Graphical model of the example problem.}
+  \label{fig:graphical_model}
+\end{figure}
+In order to demonstrate BayesPy, this section solves an extremely simple problem
+but which includes the main steps of using BayesPy.  The task is to estimate the
+unknown mean and precision parameters of a Gaussian distribution given ten
+observations.  Thus, the likelihood function is Gaussian:
+\begin{align}
+  p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
+\end{align}
+where the Gaussian distribution is parameterized by precision instead of
+variance.  The unknown mean $\mu$ and precision $\tau$ are given broad Gaussian
+and gamma priors, respectively:
+\begin{align}
+  p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}) & p(\tau) &= \mathcal{G}(\tau|10^{-6},
+  10^{-6})
+\end{align}
+Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
+The model is constructed in BayesPy by creating nodes:
+\begin{lstlisting}
+from bayespy.nodes import GaussianARD, Gamma
+mu = GaussianARD(0, 1e-6)
+tau = Gamma(1e-6, 1e-6)
+\end{lstlisting}
+These two nodes represent the two unknown variables that we are interested in.
+They are the parameters of the Gaussian distribution of the observed variable:
+\begin{lstlisting}
+y = GaussianARD(mu, tau, plates=(10,))
+\end{lstlisting}
+The number of observations is given as plates which define the number of
+repetitions.  Note that the syntax of the model construction is similar to the
+mathematical formulation which defines the conditional distributions for each
+variable.  Thus, it is easy to understand the model from the Python code.  After
+the model has been constructed, data is provided by observing nodes:
+\begin{lstlisting}
+y.observe([4.5, 3.9, 6.3, 5.6, 4.9, 2.8, 7.4, 6.1, 4.8, 2.1])
+\end{lstlisting}
+Next, we want to find the posterior approximation for our latent variables.  We
+create the VB inference engine:
+\begin{lstlisting}
+from bayespy.inference import VB
+Q = VB(y, mu, tau)
+\end{lstlisting}
+The inference engine takes all the stochastic nodes of the model as input.  The
+message passing algorithm can be run by calling \texttt{update} method:
+\begin{lstlisting}
+Q.update(repeat=100)
+\end{lstlisting}
+The algorithm is run for 100 iterations or until it converges.  In this case,
+the algorithm converges in four iterations.  The posterior approximation can be
+examined, for instance, by plotting the probability density functions of the
+latent variables:
+\begin{lstlisting}
+import bayespy.plot as bpplt
+import numpy as np
+bpplt.pyplot.figure()
+bpplt.pdf(Q[mu], np.linspace(0, 10, num=300))
+bpplt.pyplot.figure()
+bpplt.pdf(Q[tau], np.linspace(0, 2, num=300))
+bpplt.pyplot.show()
+\end{lstlisting}
+Note that the standard \texttt{pyplot} module of Matplotlib is available at the
+\texttt{plot} module of BayesPy.  Figure~\ref{fig:posterior} shows the resulting
+approximate posterior distributions.
+\begin{figure}
+  \begin{center}
+    \begin{subfigure}[b]{0.4\textwidth}
+      \includegraphics[width=\textwidth]{fig_mu}
+      \caption{$q(\mu)$}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.4\textwidth}
+      \includegraphics[width=\textwidth]{fig_tau}
+      \caption{$q(\tau)$}
+    \end{subfigure}
+  \end{center}
+  \caption{The posterior approximations of $\mu$ and $\tau$.}
+  \label{fig:posterior}
+\end{figure}
+
+\section{Conclusion and Future Work}
+
+BayesPy provides a simple and efficient way to construct conjugate exponential
+family models and to find the variational Bayesian posterior approximation in
+Python.
+%
+Future plans for BayesPy include implementing more inference engines (e.g.,
+maximum likelihood, expectation propagation and Gibbs sampling), improving the
+VB engine (e.g., collapsed variational inference \citep{Hensman:2012} and
+Riemannian conjugate gradient method \citep{Honkela:2010}) and implementing nodes
+for non-parametric models (e.g., Gaussian processes and Dirichlet processes).
 
@@
 
-\acks{?}
+%\acks{?}
 
 \vskip 0.2in
-\bibliography{bibliography.bib}
+\bibliography{bibliography/bibliography.bib}
 
 \end{document}
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-PDF-mode: t
+%%% TeX-master: t
+%%% End:

### 2014-08-14 bayespy.tex
Fix keywords

@@
   BayesPy is an open-source Python software package for performing variational
-  Bayesian inference.  It supports conjugate exponential family models by
-  utilizing the variational message passing framework.  By removing the tedious
+  Bayesian inference.  It is based on the variational message passing framework
+  and supports conjugate exponential family models.  By removing the tedious
   task of implementing the variational Bayesian update equations, the user can
@@
 \begin{keywords}
-  variational Bayes
+  Variational Bayes, variational message passing, Python, probabilistic
+  programming
 \end{keywords}
@@
 observations.  Thus, the likelihood function is Gaussian:
-\begin{align}
+\begin{align*}
   p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
-\end{align}
+\end{align*}
 where the Gaussian distribution is parameterized by precision instead of
@@
 and gamma priors, respectively:
-\begin{align}
+\begin{align*}
   p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}) & p(\tau) &= \mathcal{G}(\tau|10^{-6},
   10^{-6})
-\end{align}
+\end{align*}
 Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
@@
 variable.  Thus, it is easy to understand the model from the Python code.  After
-the model has been constructed, data is provided by observing nodes:
+the model has been constructed, some nodes are marked as observed and given
+data:
 \begin{lstlisting}
@@
 
-\section{Conclusion and Future Work}
+\section{Conclusions}
 

### 2014-08-14 bayespy.tex
Minor fine tuning

@@
 \end{lstlisting}
-Note that the standard \texttt{pyplot} module of Matplotlib is available at the
-\texttt{plot} module of BayesPy.  Figure~\ref{fig:posterior} shows the resulting
-approximate posterior distributions.
+Figure~\ref{fig:posterior} shows the resulting approximate posterior
+distributions.  The module \texttt{bayespy.plot} contains functions for simple
+plotting tasks, such as plotting the probability density function or the Hinton
+diagram of a variable.  Note that the standard \texttt{pyplot} module of
+Matplotlib is available at the \texttt{plot} module of BayesPy.
 \begin{figure}

### 2014-08-14 cover.tex
Add cover letter

\documentclass{article}

\usepackage{url}

\title{Cover Letter to JMLR Machine Learning Open Source Software Submission:\\
BayesPy: Variational Bayesian Inference in Python}

\author{Jaakko Luttinen\\
  \url{jaakko.luttinen@aalto.fi}\\
    Aalto University, Finland}

\begin{document}

\maketitle

\begin{itemize}

\item There are no previous publications by the author that significantly
  overlap with this submission.

\item There are no co-authors to be aware of the current submission and consent
  to its review by JMLR.

\item Possible conflict of interest with the following persons: Antti Honkela,
  Samuel Kaski

\item Suggestions of action editors: Geoffrey Holmes, Cheng Soon Ong, Mark Reid

\end{itemize}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-PDF-mode: t
%%% TeX-master: t
%%% End:

### 2014-11-05 experiments/pca_infernet.cs
Add initial Infer.NET scripts

@@ -1,0 +1,341 @@
+﻿using System;
+using System.Text;
+using MicrosoftResearch.Infer;
+using MicrosoftResearch.Infer.Distributions;
+using MicrosoftResearch.Infer.Maths;
+using MicrosoftResearch.Infer.Models;
+
+namespace MicrosoftResearch.Infer.Tutorials
+{
+	/// <summary>
+	/// Bayesian PCA Model
+	/// </summary>
+	public class BayesianPCAModel
+	{
+		// Inference engine
+		public InferenceEngine engine = null;
+		// Model variables
+		public Variable<int> vN = null;
+		public Variable<int> vD = null;
+		public Variable<int> vM = null;
+		public VariableArray2D<double> vData = null;
+		public VariableArray<Vector> vW = null;
+		public VariableArray<Vector> vZ = null;
+		public VariableArray2D<double> vT = null;
+		public VariableArray2D<double> vU = null;
+		public VariableArray<double> vMu = null;
+		public VariableArray<double> vPi = null;
+		public VariableArray<double> vAlpha = null;
+		// Priors - these are declared as distribution variables
+		// so that we can set them at run-time. They are variables
+		// from the perspective of the 'Random' factor which takes
+		// a distribution as an argument.
+		public Variable<Gamma> priorAlpha = null;
+		public Variable<Gaussian> priorMu = null;
+		public Variable<Gamma> priorPi = null;
+		// Model ranges
+		public Range rN = null;
+		public Range rD = null;
+		public Range rM = null;
+
+		/// <summary>
+		/// Model constructor
+		/// </summary>
+		public BayesianPCAModel()
+		{
+			// The various dimensions will be set externally...
+			vN = Variable.New<int>().Named("NumObs");
+			vD = Variable.New<int>().Named("NumFeats");
+			vM = Variable.New<int>().Named("MaxComponents");
+			rN = new Range(vN).Named("N");
+			rD = new Range(vD).Named("D");
+			rM = new Range(vM).Named("M");
+			// ... as will the data
+			vData = Variable.Array<double>(rN, rD).Named("data");
+			// ... and the priors
+			priorAlpha = Variable.New<Gamma>().Named("PriorAlpha");
+			priorMu = Variable.New<Gaussian>().Named("PriorMu");
+			priorPi = Variable.New<Gamma>().Named("PriorPi");
[…]

### 2015-03-09 experiments/make_plots.py
Add results and ready experiment functions

@@ -1,0 +1,33 @@
+# Copyright (c) 2015 Jaakko Luttinen
+
+from utils import plot
+
+import matplotlib.pyplot as plt
+
+if __name__ == "__main__":
+
+    plt.figure()
+    plot("pca", 1, maxiter=200)
+    plt.xlim(0, 6)
+    plt.ylim(-7500, -4500)
+    plt.savefig("fig_pca_01.pdf", frameon=False)
+    
+    plt.figure()
+    plot("pca", 2, maxiter=200)
+    plt.xlim(0, 70)
+    plt.ylim(-120000, -70000)
+    plt.savefig("fig_pca_02.pdf", frameon=False)
+    
+    plt.figure()
+    plot("mog", 1, maxiter=200)
+    plt.xlim(0, 3)
+    plt.ylim(-1000, -750)
+    plt.savefig("fig_mog_01.pdf", frameon=False)
+
+    plt.figure()
+    plot("mog", 2, maxiter=200)
+    plt.xlim(0, 150)
+    plt.ylim(-40000, -32000)
+    plt.savefig("fig_mog_02.pdf", frameon=False)
+
+    plt.show()

### 2015-03-09 experiments/mog_bayespy.py

@@ -42,7 +42,7 @@ def mog_model(N, K, D):
     # Construct the Gaussian mixture model
 
     # K prior weights (for components)
-    alpha = nodes.Dirichlet(1.0*np.ones(K),
+    alpha = nodes.Dirichlet(0.1*np.ones(K),
                             name='alpha')
     # N K-dimensional cluster assignments (for data)
     z = nodes.Categorical(alpha,
@@ -62,19 +62,19 @@ def mog_model(N, K, D):
     # Break symmetry
     z.initialize_from_random()
 
-    return VB(Y, mu, Lambda, z, alpha)
+    return VB(Y, mu, z, Lambda, alpha)
 
 
-def generate_data(N, D, K, seed=1):
+def generate_data(N, D, K, seed=1, spread=3):
     """
     Generate data from a mixture of Gaussians model
     """
 
     np.random.seed(seed)
 
-    mu = 3*np.random.randn(K, D)
+    mu = spread*np.random.randn(K, D)
     # Lambda is actually precision matrix (inverse covariance)
-    Lambda = random.covariance(D, size=K)
+    Lambda = random.covariance(D, size=K, nu=2*D)
     pi = random.dirichlet(5*np.ones(K))
 
     y = np.zeros((N,D))
@@ -85,9 +85,8 @@ def generate_data(N, D, K, seed=1):
 
     np.savetxt('mog-data-%02d.csv' % seed, y, delimiter=',', fmt='%f')
 
-    print(mu)
-    plt.plot(y[:,0], y[:,1], 'rx')
-    plt.show()
+    #plt.plot(y[:,0], y[:,1], 'rx')
+    #plt.show()
 
     return y
 
@@ -99,11 +98,11 @@ def plot(seed=1, maxiter=None):
     utils.plot('mog', seed, maxiter=maxiter)
 
 
-def run(N=2000, D=3, K=20, seed=1, maxiter=200):
+def run(N=2000, D=3, K=20, seed=1, maxiter=200, spread=3.0):
 
     # Get data
     print("Generating data...")
-    y = generate_data(N, D, np.ceil(K/2), seed=seed)
+    y = generate_data(N, D, int(np.ceil(K/2)), seed=seed, spread=spread)
 
     # Construct model
     Q = mog_model(N, K, D)
@@ -113,10 +112,10 @@ def run(N=2000, D=3, K=20, seed=1, maxiter=200):
 
     # Run inference
[…]

### 2015-03-09 experiments/mog_infernet.cs

@@ -48,19 +48,25 @@ namespace MicrosoftResearch.Infer.Tutorials
                 // Mixture component means
                 means = Variable.Array<Vector>(k).Named("means");			
                 means[k] = Variable.VectorGaussianFromMeanAndPrecision(
-                    Vector.Zero(d.SizeAsInt),
-                    PositiveDefiniteMatrix.IdentityScaledBy(d.SizeAsInt,0.01)).ForEach(k);
+                    Vector.Zero(D),
+                    PositiveDefiniteMatrix.IdentityScaledBy(D, 0.01)
+                ).ForEach(k);
 	
                 // Mixture component precisions
+                // NOTE: Compared to BayesPy, prior parameters are multiplied by
+                // 0.5 because of different parameterization of the Wishart
+                // distribution.
                 precs = Variable.Array<PositiveDefiniteMatrix>(k).Named("precs");
-                precs[k] = Variable.WishartFromShapeAndRate(d.SizeAsInt,
-                                                            PositiveDefiniteMatrix.IdentityScaledBy(d.SizeAsInt,d.SizeAsInt)).ForEach(k);
+                precs[k] = Variable.WishartFromShapeAndRate(
+                    0.5*D,
+                    PositiveDefiniteMatrix.IdentityScaledBy(D, 0.5*D)
+                ).ForEach(k);
 			
                 // Mixture weights
                 double[] a = new double[K];
                 for (int i = 0; i < K; i++)
                 {
-                    a[i] = 1;
+                    a[i] = 0.1;
                 }
                 weights = Variable.Dirichlet(k, a).Named("weights");	
 
@@ -131,7 +137,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                 mog.z.InitialiseTo(Distribution<int>.Array(zinit));
 
                 // The inference
-                //Process proc = Process.GetCurrentProcess();
                 double logEvidence = 0;
                 var loglike = new List<double>();
                 var cputime = new List<double>();
@@ -161,8 +166,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                 }
 
                 Console.WriteLine("Dist over pi=" + mog.engine.Infer(mog.weights));
-                //Console.WriteLine("Dist over means=\n" + mog.ie.Infer(mog.means));
-                //Console.WriteLine("Dist over precs=\n" + mog.ie.Infer(mog.precs));
 
                 SaveResults(string.Format("mog-results-{0:00}-infernet.csv", seed),
                             loglike.ToArray(), cputime.ToArray());

### 2015-03-09 experiments/pca_bayespy.py

@@ -39,7 +39,7 @@ import bayespy.plot as bpplt
 def model(M, N, D):
     # Construct the PCA model with ARD
 
-    if False:
+    if True:
         # ARD covariance
         alpha = nodes.Gamma(1e-2,
                             1e-2,
@@ -104,9 +104,15 @@ def model(M, N, D):
 
 def generate_data(M, N, D, seed=1):
     np.random.seed(seed)
-    W = np.random.randn(M, D)
+    #U = np.linalg.qr(np.random.randn(M,M))[0]
+    #V = np.linalg.qr(np.random.randn(N,N))[0]
+    #S = np.eye(M, N)
+    #S[:D,:D] *= np.linspace(8, 8, D) #(1 + np.arange(D))
+    #y = np.dot(U, np.dot(S, V))
+    W = np.random.randn(M, D) / D
     X = np.random.randn(D, N)
-    y = np.dot(W, X) + 0.5*np.random.randn(M, N)
+    y = np.dot(W, X) + 0.3*np.random.randn(M, N)
+    (u,s,v) = np.linalg.svd(y)
     np.savetxt('pca-data-%02d.csv' % seed, y, delimiter=',', fmt='%f')
     return y
 
@@ -143,7 +149,7 @@ def run(M=100, N=1000, D=10, seed=42, rotate=False, maxiter=200, debug=False):
     Q.update(repeat=maxiter, tol=0)
 
     v = np.array([Q.L[:(Q.iter+1)], Q.cputime[:(Q.iter+1)]]).T
-    np.savetxt("mog-results-%02d-bayespy.csv" % seed, v, delimiter=",")
+    np.savetxt("pca-results-%02d-bayespy.csv" % seed, v, delimiter=",")
 
     print(Q['tau'])
 

### 2015-03-09 experiments/pca_infernet.cs

@@ -302,6 +302,7 @@ namespace MicrosoftResearch.Infer.Tutorials
                     cputime.Add((endtime - starttime) / 1000.0);
                     // Compute lower bound
                     logEvidence = bpca.bound(); //bpca.engine.Infer<Bernoulli>(bpca.evidence).LogOdds;
+                    loglike.Add(logEvidence);
                     // Print progress
                     Console.WriteLine("Iteration {0}: loglike={1} ({2} ms)",
                                       j+1,

### 2015-03-09 experiments/run_experiments.sh

@@ -1,13 +1,16 @@
 #!/bin/bash
 
+# PCA experiments
+python pca_bayespy.py --n=500 --m=20 --d=10 --maxiter=200 --seed=1
+# make && mono pca_infernet.exe 10 1 200 0
+make && mono pca_infernet.exe 10 1 200 1
+python pca_bayespy.py --n=2000 --m=100 --d=40 --maxiter=200 --seed=2
+# make && mono pca_infernet.exe 40 2 200 0
+make && mono pca_infernet.exe 40 2 200 1
+
 # MoG experiments
-python mog_bayespy.py --n=200 --k=10 --d=2 --maxiter=200 --seed=1
+python mog_bayespy.py --n=200 --k=10 --d=2 --maxiter=200 --seed=1 --spread=3.0
 make && mono mog_infernet.exe 10 1 200
-python mog_bayespy.py --n=2000 --k=40 --d=5 --maxiter=200 --seed=2
+python mog_bayespy.py --n=2000 --k=40 --d=10 --maxiter=200 --seed=2 --spread=1.0
 make && mono mog_infernet.exe 40 2 200
 
-# PCA experiments
-python pca_bayespy.py --n=200 --m=20 --d=10 --maxiter=200 --seed=1
-make && mono pca_infernet.exe 10 1 200
-python pca_bayespy.py --n=2000 --m=200 --d=40 --maxiter=200 --seed=2
-make && mono pca_infernet.exe 40 2 200

### 2015-03-10 response.tex
Add response skeleton

\documentclass{article}

\usepackage{url}

\title{Review Reply to JMLR Machine Learning Open Source Software Submission (14-324):\\
BayesPy: Variational Bayesian Inference in Python}

\author{Jaakko Luttinen\\
  \url{jaakko.luttinen@aalto.fi}\\
    Aalto University, Finland}

\begin{document}

\maketitle

\section{Responses to the Reviewers of Submission 14-324 (v1)}

I would like to thank..

added features etc

% \subsection{Decision (meta-review)}

% The blah blah

% \begin{itemize}

% \end{itemize}

\subsection{Reviewer 1}

\begin{itemize}

\item \underline{syntax}\\
  yeah yeah

\item \underline{speed comparison}\\
  ooh yeeeeah

\item \underline{supported models}\\

\item \underline{scheduling}\\

\end{itemize}

\subsection{Reviewer 2}

\begin{itemize}

\item \underline{installation instructions}\\

\item \underline{user community}\\

\item \underline{speed comparison}

\end{itemize}

\subsection{Reviewer 3}

Reviewer 3 did not point out any BLAH BLAH BLAH

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-PDF-mode: t
%%% TeX-master: t
%%% End:

### 2015-03-10 bayespy.tex
Fix punctuation
Update bibliography
Add advanced VB methods paragraph

@@
 
-\section{Design and Features}
+\section{Features}
 
@@
 intuitive to read and write by making it similar to the mathematical formulation
-of the model.  Other notable features of BayesPy include easy handling of
-missing values, parameter expansions \citep{Qi:2007} for speeding up inference
-and monitoring of the variables by plotting the posterior distributions during
-the message passing iteration.  For developers, the unit testing framework helps
-in finding bugs, making changes and implementing new features in a robust
-manner.
+of the model.  Missing values are easy to handle and the variables can be
+monitored by plotting the posterior distributions during the learning.  BayesPy
+has also preliminary support for some advanced VB methods such as stochastic
+variational inference \citep{Hoffman:2013}, deterministic annealing
+\citep{Katahira:2008}, collapsed inference \citep{Hensman:2012},
+Riemannian-gradient-based optimization \citep{Honkela:2010}, parameter
+expansions \citep{Qi:2007} and pattern searches \citep{Honkela:2003}.  For
+developers, the unit testing framework helps in finding bugs, making changes and
+implementing new features in a robust manner.
 
@@
 \begin{align*}
-  p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}) & p(\tau) &= \mathcal{G}(\tau|10^{-6},
-  10^{-6})
+  p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}), & p(\tau) &= \mathcal{G}(\tau|10^{-6},
+  10^{-6}).
 \end{align*}

### 2015-03-11 bayespy.tex
Add BayesPy vs Infer.NET comparison

@@
 \firstpageno{1}
+
+\lstset{ %
+  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
+  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+  breaklines=true,                 % sets automatic line breaking
+  captionpos=b,                    % sets the caption-position to bottom
+  commentstyle=\color{green},    % comment style
+  deletekeywords={...},            % if you want to delete keywords from the given language
+  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
+  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
+%  frame=single,                    % adds a frame around the code
+  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+  keywordstyle=\color{blue},       % keyword style
+  morekeywords={*,...},            % if you want to add more keywords to the set
+  firstnumber=last,
+  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
+  numbersep=10pt,                   % how far the line-numbers are from the code
+  numberstyle=\tiny\color{black}, % the style that is used for the line-numbers
+  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
+  showstringspaces=false,          % underline spaces within strings only
+  showtabs=false,                  % show tabs within strings adding particular underscores
+  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
+  stringstyle=\color{orange},     % string literal style
+  tabsize=2,                       % sets default tabsize to 2 spaces
+%  title=\lstname,                   % show the filename of files included with \lstinputlisting; also try caption instead of title
+  belowcaptionskip=1\baselineskip,
+  breaklines=true,
+  frame=L,
+  xleftmargin=1.3\parindent,
+  showstringspaces=false,
+  basicstyle=\footnotesize\ttfamily,
+%  keywordstyle=\bfseries\color{green!40!black},
+%  commentstyle=\itshape\color{purple!40!black},
+%  identifierstyle=\color{blue},
+%  stringstyle=\color{orange},
+  language=Python
+}
 
@@
 
-BayesPy can be used to construct a wide range of conjugate exponential family
-models.  The documentation provides detailed examples of how to construct, for
-instance, regression models, principal component analysis models, linear
-state-space models, mixture models and hidden Markov models.  In addition,
-BayesPy has been used in two publications about parameter expansion and
-time-varying dynamics for linear state-space models
-\citep{Luttinen:2013,Luttinen:2014}, which show that it is applicable to large
-complex models requiring efficient inference.
+BayesPy can be used to construct conjugate exponential family models.  The
+documentation provides detailed examples of how to construct, for instance,
+regression models, principal component analysis models, linear state-space
+models, mixture models and hidden Markov models.  In addition, BayesPy has been
+used in two publications about parameter expansion and time-varying dynamics for
+linear state-space models \citep{Luttinen:2013,Luttinen:2014}, which show that
+it is applicable to large complex models requiring efficient inference.
 
@@
 
-\lstset{ %
-  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
-  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
-  breaklines=true,                 % sets automatic line breaking
-  captionpos=b,                    % sets the caption-position to bottom
-  commentstyle=\color{green},    % comment style
-  deletekeywords={...},            % if you want to delete keywords from the given language
-  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
-  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
-%  frame=single,                    % adds a frame around the code
-  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
-  keywordstyle=\color{blue},       % keyword style
-  morekeywords={*,...},            % if you want to add more keywords to the set
-  firstnumber=last,
-  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
-  numbersep=10pt,                   % how far the line-numbers are from the code
-  numberstyle=\tiny\color{black}, % the style that is used for the line-numbers
-  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
-  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
-  showstringspaces=false,          % underline spaces within strings only
-  showtabs=false,                  % show tabs within strings adding particular underscores
-  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
-  stringstyle=\color{orange},     % string literal style
-  tabsize=2,                       % sets default tabsize to 2 spaces
-%  title=\lstname,                   % show the filename of files included with \lstinputlisting; also try caption instead of title
-  belowcaptionskip=1\baselineskip,
-  breaklines=true,
-  frame=L,
-  xleftmargin=1.3\parindent,
-  showstringspaces=false,
-  basicstyle=\footnotesize\ttfamily,
-%  keywordstyle=\bfseries\color{green!40!black},
-%  commentstyle=\itshape\color{purple!40!black},
-%  identifierstyle=\color{blue},
-%  stringstyle=\color{orange},
-  language=Python
-}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node [obs] (yn) {$y_n$};
-    \node [latent, above left=of yn] (mu) {$\mu$};
-    \node [latent, above right=of yn] (tau) {$\tau$};
-
-    \edge {mu, tau} {yn};
-
-    \tikzstyle{plate caption} = [caption, node distance=0, inner sep=0pt,
-    below=5pt of #1.south]
-
-    \plate {} {(yn)} {$n=1,\ldots,10$};
-  \end{tikzpicture}
-  \caption{Graphical model of the example problem.}
-  \label{fig:graphical_model}
-\end{figure}
-In order to demonstrate BayesPy, this section solves an extremely simple problem
-but which includes the main steps of using BayesPy.  The task is to estimate the
-unknown mean and precision parameters of a Gaussian distribution given ten
-observations.  Thus, the likelihood function is Gaussian:
-\begin{align*}
-  p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
-\end{align*}
-where the Gaussian distribution is parameterized by precision instead of
-variance.  The unknown mean $\mu$ and precision $\tau$ are given broad Gaussian
-and gamma priors, respectively:
-\begin{align*}
-  p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}), & p(\tau) &= \mathcal{G}(\tau|10^{-6},
-  10^{-6}).
-\end{align*}
-Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
-The model is constructed in BayesPy by creating nodes:
+This section demonstrates BayesPy with a Gaussian mixture model.  A simple
+artificial dataset is created by drawing 200 samples from a 2-dimensional
+Gaussian distribution.  80 samples have mean $[1, 1]$ and 120 samples have mean
+$[0, 0]$:
 \begin{lstlisting}
-from bayespy.nodes import GaussianARD, Gamma
-mu = GaussianARD(0, 1e-6)
-tau = Gamma(1e-6, 1e-6)
+import numpy as np
+N = 200, D = 2
+data = np.random.randn(N, D)
+data[:80,:] += np.ones(D)
 \end{lstlisting}
-These two nodes represent the two unknown variables that we are interested in.
-They are the parameters of the Gaussian distribution of the observed variable:
+We construct a mixture model for the data and assume that the parameters and the
+cluster assignments are unknown, even the true number of clusters.  The model
+uses a maximum number of five clusters but the effective number of clusters will
+be determined automatically:
 \begin{lstlisting}
-y = GaussianARD(mu, tau, plates=(10,))
+K = 5
 \end{lstlisting}
-The number of observations is given as plates which define the number of
-repetitions.  Note that the syntax of the model construction is similar to the
-mathematical formulation which defines the conditional distributions for each
-variable.  Thus, it is easy to understand the model from the Python code.  After
-the model has been constructed, some nodes are marked as observed and given
-data:
+The unknown cluster means and precision matrices are given Gaussian and Wishart
+prior distributions:
 \begin{lstlisting}
-y.observe([4.5, 3.9, 6.3, 5.6, 4.9, 2.8, 7.4, 6.1, 4.8, 2.1])
+from bayespy import nodes
+mu = nodes.Gaussian(np.zeros(D), 0.01*np.identity(D), plates=(K,))
+Lambda = nodes.Wishart(D, 0.01*np.identity(D), plates=(K,))
+\end{lstlisting}
+The \texttt{plates} keyword argument is used to define repetitions similarly as
+the plate notation in graphical models.  The cluster assignments are categorical
+variables, and the cluster probabilities are given a Dirichlet prior
+distribution:
+\begin{lstlisting}
+alpha = nodes.Dirichlet(0.01*np.ones(K))
+z = nodes.Categorical(alpha, plates=(N,))
+\end{lstlisting}
+The observations are from a Gaussian mixture distribution:
+\begin{lstlisting}
+y = nodes.Mixture(Z, nodes.Gaussian, mu, Lambda)
+\end{lstlisting}
+The second argument for the \texttt{Mixture} node defines the type of the
+mixture distribution, in this case Gaussian.  The variable is marked as observed
+by providing the data:
+\begin{lstlisting}
+y.observe(data)
 \end{lstlisting}
 Next, we want to find the posterior approximation for our latent variables.  We
-create the VB inference engine:
+create the variational Bayesian inference engine:
 \begin{lstlisting}
 from bayespy.inference import VB
-Q = VB(y, mu, tau)
+Q = VB(y, mu, z, Lambda, alpha)
 \end{lstlisting}
-The inference engine takes all the stochastic nodes of the model as input.  The
-message passing algorithm can be run by calling \texttt{update} method:
+Before running the VMP algorithm, the symmetry in the model is broken by
+a random initialization of the cluster assignments:
+\begin{lstlisting}
+z.initialize_from_random()
+\end{lstlisting}
+Without the random initialization, the clusters would not be separated.  The VMP
+algorithm is run for 100 iterations or until convergence:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The algorithm is run for 100 iterations or until it converges.  In this case,
-the algorithm converges in four iterations.  The posterior approximation can be
-examined, for instance, by plotting the probability density functions of the
-latent variables:
+The results can be examined, for instance, by printing the posterior
+distributions of some variables:
 \begin{lstlisting}
-import bayespy.plot as bpplt
-import numpy as np
-bpplt.pyplot.figure()
-bpplt.pdf(Q[mu], np.linspace(0, 10, num=300))
-bpplt.pyplot.figure()
-bpplt.pdf(Q[tau], np.linspace(0, 2, num=300))
-bpplt.pyplot.show()
+print(alpha)
+print(mu)
 \end{lstlisting}
-Figure~\ref{fig:posterior} shows the resulting approximate posterior
-distributions.  The module \texttt{bayespy.plot} contains functions for simple
-plotting tasks, such as plotting the probability density function or the Hinton
-diagram of a variable.  Note that the standard \texttt{pyplot} module of
-Matplotlib is available at the \texttt{plot} module of BayesPy.
-\begin{figure}
-  \begin{center}
-    \begin{subfigure}[b]{0.4\textwidth}
-      \includegraphics[width=\textwidth]{fig_mu}
-      \caption{$q(\mu)$}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.4\textwidth}
-      \includegraphics[width=\textwidth]{fig_tau}
-      \caption{$q(\tau)$}
-    \end{subfigure}
-  \end{center}
-  \caption{The posterior approximations of $\mu$ and $\tau$.}
-  \label{fig:posterior}
-\end{figure}
+In addition, module \texttt{bayespy.plot} contains several plotting functions to
+examine the results visually.
+
+\section{Comparison with Infer.NET}
+
+This section provides a brief comparison with Infer.NET because it is another
+active project implementing the variational message passing algorithm, whereas
+the other related active projects implement MCMC.  The main advantages of
+Infer.NET over BayesPy are its support for non-conjugate models and other
+inference engines.  On the other hand, BayesPy is open-source software and
+supports several advanced VB methods (e.g., stochastic variational inference and
+collapsed inference).
+
+The speed of the packages were compared by using two widely used models: a
+Gaussian mixture model (GMM) and principal component analysis
+(PCA).\footnote{The scripts for running the experiments are available in the
+  supplementary material.}  Both models were run for small and large
+artificially generated datasets.  For GMM, the small model used 10 clusters for
+200 observations with 2 dimensions, and the large model used 40 clusters for
+2000 observations with 10 dimensions.  For PCA, the small model used
+10-dimensional latent space for 500 observations with 20 dimensions, and the
+large model used 40-dimensional latent space for 2000 observations with 100
+dimensions.  Infer.NET used a fully factorizing approximation for the PCA model,
+whereas BayesPy did not factorize with respect to the latent space dimensions.
+The experiments were run on a Linux computer with a quad-core processor
+(i7-4702MQ).  Because the packages run practically identical algorithms and thus
+converged to similar solutions in a similar number of iterations (50--100
+iterations depending on the dataset), we compare the average CPU time per
+iteration.
+
+The results are summarized in Table~\ref{tab:speed}.  On the smaller datasets,
+Infer.NET is slightly faster because BayesPy has some overhead from using
+Python.  However, on the larger datasets, BayesPy is significantly faster
+because it uses BLAS and LAPACK libraries for fast numerical
+operations.\footnote{NumPy must use BLAS/LAPACK because BayesPy uses NumPy for
+  numerical operations.}  In addition, if using multi-threaded BLAS/LAPACK, the
+difference in wall-clock time can be even larger than in CPU time.  The speed of
+BayesPy for these models could potentially be improved by using collapsed or
+stochastic variational inference, and the accuracy could be improved by using
+Gaussian-Wishart nodes which do not factorize with respect to the Gaussian and
+Wishart variables.
+
+\begin{table}[tb]
+  \centering
+  \caption{The average CPU time (in milliseconds) per iteration for each dataset.}
+  \small
+  \begin{tabular}{ccccc}
+    &
+    Small GMM
+    &
+    Large GMM
+    &
+    Small PCA
+    &
+    Large PCA
+    \\
+    \hline
+    BayesPy     & 45 & 210  & 52 & 380
+    \\
+    Infer.NET   & 26 & 4400 & 37 & 2200
+  \end{tabular}
+  \label{tab:speed}
+\end{table}
+
+% related active software projects discussed in the introduction,
+
+% Because the other
+% active Bayesian software packages discussed in the introduction do not implement
+% the variational message passing algorithm,
+
+%* compiled
+
+%* non-conjugate
+
+%* other engines
+
+%* open source
+
+%* advanced
+
+%* dynamic nodes
+
+% \begin{figure}
+%   \centering
+%   \begin{tikzpicture}
+%     \node [obs] (yn) {$y_n$};
+%     \node [latent, above left=of yn] (mu) {$\mu$};
+%     \node [latent, above right=of yn] (tau) {$\tau$};
+
+%     \edge {mu, tau} {yn};
+
+%     \tikzstyle{plate caption} = [caption, node distance=0, inner sep=0pt,
+%     below=5pt of #1.south]
+
+%     \plate {} {(yn)} {$n=1,\ldots,10$};
+%   \end{tikzpicture}
+%   \caption{Graphical model of the example problem.}
+%   \label{fig:graphical_model}
+% \end{figure}
+% In order to demonstrate BayesPy, this section solves an extremely simple problem
+% but which includes the main steps of using BayesPy.  The task is to estimate the
+% unknown mean and precision parameters of a Gaussian distribution given ten
+% observations.  Thus, the likelihood function is Gaussian:
+% \begin{align*}
+%   p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
+% \end{align*}
+% where the Gaussian distribution is parameterized by precision instead of
+% variance.  The unknown mean $\mu$ and precision $\tau$ are given broad Gaussian
+% and gamma priors, respectively:
+% \begin{align*}
+%   p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}), & p(\tau) &= \mathcal{G}(\tau|10^{-6},
+%   10^{-6}).
+% \end{align*}
+% Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
+% The model is constructed in BayesPy by creating nodes:
+% \begin{lstlisting}
+% from bayespy.nodes import GaussianARD, Gamma
+% mu = GaussianARD(0, 1e-6)
+% tau = Gamma(1e-6, 1e-6)
+% \end{lstlisting}
+% These two nodes represent the two unknown variables that we are interested in.
+% They are the parameters of the Gaussian distribution of the observed variable:
+% \begin{lstlisting}
+% y = GaussianARD(mu, tau, plates=(10,))
+% \end{lstlisting}
+% The number of observations is given as plates which define the number of
+% repetitions.  Note that the syntax of the model construction is similar to the
+% mathematical formulation which defines the conditional distributions for each
+% variable.  Thus, it is easy to understand the model from the Python code.  After
+% the model has been constructed, some nodes are marked as observed and given
+% data:
+% \begin{lstlisting}
+% y.observe([4.5, 3.9, 6.3, 5.6, 4.9, 2.8, 7.4, 6.1, 4.8, 2.1])
+% \end{lstlisting}
+% Next, we want to find the posterior approximation for our latent variables.  We
+% create the VB inference engine:
+% \begin{lstlisting}
+% from bayespy.inference import VB
+% Q = VB(y, mu, tau)
+% \end{lstlisting}
+% The inference engine takes all the stochastic nodes of the model as input.  The
+% message passing algorithm can be run by calling \texttt{update} method:
+% \begin{lstlisting}
+% Q.update(repeat=100)
+% \end{lstlisting}
+% The algorithm is run for 100 iterations or until it converges.  In this case,
+% the algorithm converges in four iterations.  The posterior approximation can be
+% examined, for instance, by plotting the probability density functions of the
+% latent variables:
+% \begin{lstlisting}
+% import bayespy.plot as bpplt
+% import numpy as np
+% bpplt.pyplot.figure()
+% bpplt.pdf(Q[mu], np.linspace(0, 10, num=300))
+% bpplt.pyplot.figure()
+% bpplt.pdf(Q[tau], np.linspace(0, 2, num=300))
+% bpplt.pyplot.show()
+% \end{lstlisting}
+% Figure~\ref{fig:posterior} shows the resulting approximate posterior
+% distributions.  The module \texttt{bayespy.plot} contains functions for simple
+% plotting tasks, such as plotting the probability density function or the Hinton
+% diagram of a variable.  Note that the standard \texttt{pyplot} module of
+% Matplotlib is available at the \texttt{plot} module of BayesPy.
+% \begin{figure}
+%   \begin{center}
+%     \begin{subfigure}[b]{0.4\textwidth}
+%       \includegraphics[width=\textwidth]{fig_mu}
+%       \caption{$q(\mu)$}
+%     \end{subfigure}
+%     \begin{subfigure}[b]{0.4\textwidth}
+%       \includegraphics[width=\textwidth]{fig_tau}
+%       \caption{$q(\tau)$}
+%     \end{subfigure}
+%   \end{center}
+%   \caption{The posterior approximations of $\mu$ and $\tau$.}
+%   \label{fig:posterior}
+% \end{figure}
 

### 2015-03-11 experiments/utils.py

@@ -34,4 +34,13 @@ def plot(method, seed, maxiter=None):
 
     plt.legend(['Infer.NET', 'BayesPy'], loc='lower right')
 
+    print("Method={0}, seed={1}, package={2}, CPU time={3}".format(method,
+                                                                   seed,
+                                                                   "bayespy",
+                                                                   np.mean(result_bp[:,1])))
+    print("Method={0}, seed={1}, package={2}, CPU time={3}".format(method,
+                                                                   seed,
+                                                                   "infernet",
+                                                                   np.mean(result_in[:,1])))
+
 

### 2015-03-11 bayespy.tex
Modify to page limits

@@
 
-\jmlrheading{?}{2014}{?-?}{?/??}{?/??}{Jaakko Luttinen}
+\jmlrheading{?}{2015}{?-?}{?/??}{?/??}{Jaakko Luttinen}
 
@@
 \author{\name Jaakko Luttinen \email jaakko.luttinen@aalto.fi \\
-       \addr Department of Information and Computer Science\\
+       \addr Department of Computer Science\\
        Aalto University, Finland}
@@
 
-Bayesian framework provides theoretically solid and consistent way to construct
-models and perform inference.  In practice, however, the inference is usually
-analytically intractable and is therefore based on approximation methods such as
-variational Bayes (VB), expectation propagation (EP) and Markov chain Monte
-Carlo (MCMC) \citep{Bishop:2006}.  Deriving and implementing the formulas for an
-approximation method is often straightforward but tedious, time consuming and
-error prone.
+Bayesian framework provides a theoretically solid and consistent way to
+construct models and perform inference.  In practice, however, the inference is
+usually analytically intractable and is therefore based on approximation methods
+such as variational Bayes (VB), expectation propagation (EP) and Markov chain
+Monte Carlo (MCMC) \citep{Bishop:2006}.  Deriving and implementing the formulas
+for an approximation method is often straightforward but tedious, time consuming
+and error prone.
 
-BayesPy is a Python package providing tools for constructing Bayesian models and
-performing variational Bayesian inference easily and efficiently.  It is based
-on variational message passing (VMP) framework which defines a simple and local
-message passing protocol \citep{Winn:2005}.  This enables implementation of small
-general modules that can be used as building blocks for large complex models.
-BayesPy offers a comprehensive collection of built-in blocks that can be used to
-build a wide range of models and a simple interface for implementing custom
-blocks.  The package is written for Python 3 and released under the GNU General
-Public License v3.0 (GPLv3).
+BayesPy is a Python 3 package providing tools for constructing models and
+performing VB inference easily and efficiently.  It is based on the variational
+message passing (VMP) framework which defines a simple message passing protocol
+\citep{Winn:2005}.  This enables implementation of small general nodes that can
+be used as building blocks for large complex models.  BayesPy offers a
+comprehensive collection of built-in nodes that can be used to build a wide
+range of models and a simple interface for implementing custom nodes.  The
+package is released under the GNU General Public License v3.0 (GPLv3).
 
@@
 \citep{Raiko:2007}, VIBES \citep{Bishop:2002} and Infer.NET \citep{Infer.NET}.
-Bayes Blocks is an open-source C++/Python package but it is limited to scalar
-Gaussian nodes, a few deterministic functions and fully factorial posterior
-approximation, thus making it very limited.  VIBES is an old software package
-for Java, released under the revised BSD license, but it is no longer actively
-developed.  VIBES has been replaced by Infer.NET, which supports a wide range of
-posterior approximations.  However, Infer.NET is partly closed source and
-licensed for non-commercial use only.  Instead of VB inference, mainly MCMC is
-supported by other projects such as PyMC \citep{PyMC}, OpenBUGS \citep{OpenBUGS},
-Dimple \citep{Dimple} and Stan \citep{Stan}.  Thus, there is a need for an open
-source and maintained VB software package.
+Bayes Blocks is an open-source C++/Python package but limited to scalar Gaussian
+nodes and a few deterministic functions, thus making it very limited.  VIBES is
+an old software package for Java, released under the revised BSD license, but it
+is no longer actively developed.  VIBES has been replaced by Infer.NET, which
+%supports a wide range of posterior approximations.  However, Infer.NET
+is partly closed source and licensed for non-commercial use only.  Instead of VB
+inference, mainly MCMC is supported by other projects such as PyMC \citep{PyMC},
+OpenBUGS \citep{OpenBUGS}, Dimple \citep{Dimple} and Stan \citep{Stan}.  Thus,
+there is a need for an open source and maintained VB software package.
 
@@
 documentation provides detailed examples of how to construct, for instance,
-regression models, principal component analysis models, linear state-space
-models, mixture models and hidden Markov models.  In addition, BayesPy has been
-used in two publications about parameter expansion and time-varying dynamics for
-linear state-space models \citep{Luttinen:2013,Luttinen:2014}, which show that
-it is applicable to large complex models requiring efficient inference.
+principal component analysis models, linear state-space models, mixture models
+and hidden Markov models.  In addition, BayesPy has been used in two
+publications about parameter expansion and time-varying dynamics for linear
+state-space models \citep{Luttinen:2013,Luttinen:2014}.
+% , which show that it is applicable to large complex models requiring efficient
+% inference.
 
@@
 Roughly, each node corresponds to a latent variable, a set of observations or a
-deterministic function.  Some of the nodes may be set as observed by providing
-data.  After that, inference engine is initialized and the message passing
-algorithm is run in order to obtain the posterior approximation.  The resulting
-posterior can be examined, for instance, by using a few built-in plotting
-functions or printing the parameters of the posterior distributions.
+deterministic function.
+%Some of the nodes may be set as observed by providing
+%data.
+The inference engine is used to run the message passing algorithm in order to
+obtain the posterior approximation.  The resulting posterior can be examined,
+for instance, by using a few built-in plotting functions or printing the
+parameters of the posterior distributions.
 
@@
 correspond to functions.  Built-in stochastic nodes include all common
-exponential family distributions (e.g., Gaussian, gamma, Dirichlet and Poisson
+exponential family distributions (e.g., Gaussian, gamma and Dirichlet
 distributions), a general mixture distribution and a few complex nodes for
 dynamic variables (e.g., discrete and Gaussian Markov chains).  Built-in
-deterministic nodes include a gating node and a general sum-product node.  In
-case a model cannot be constructed using the built-in nodes, the documentation
-provides instructions for implementing new nodes.
+deterministic nodes include a gating node and a general sum-product node.
+%The
+%documentation also provides instructions for implementing new nodes in case the
+%built-in collection is insufficient.
+In case a model cannot be constructed using the built-in nodes, the
+documentation provides instructions for implementing new nodes.
 
@@
 \end{lstlisting}
-We construct a mixture model for the data and assume that the parameters and the
-cluster assignments are unknown, even the true number of clusters.  The model
-uses a maximum number of five clusters but the effective number of clusters will
-be determined automatically:
+We construct a mixture model for the data and assume that the parameters, the
+cluster assignments and the true number of clusters are unknown.  The model uses
+a maximum number of five clusters but the effective number of clusters will be
+determined automatically:
 \begin{lstlisting}
@@
 The results are summarized in Table~\ref{tab:speed}.  On the smaller datasets,
-Infer.NET is slightly faster because BayesPy has some overhead from using
-Python.  However, on the larger datasets, BayesPy is significantly faster
-because it uses BLAS and LAPACK libraries for fast numerical
-operations.\footnote{NumPy must use BLAS/LAPACK because BayesPy uses NumPy for
-  numerical operations.}  In addition, if using multi-threaded BLAS/LAPACK, the
-difference in wall-clock time can be even larger than in CPU time.  The speed of
-BayesPy for these models could potentially be improved by using collapsed or
-stochastic variational inference, and the accuracy could be improved by using
+Infer.NET is slightly faster because BayesPy has some overhead of Python.
+However, on the larger datasets, BayesPy is significantly faster because it uses
+BLAS and LAPACK libraries for fast numerical operations.\footnote{Assuming that
+  NumPy uses BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}
+In addition, if using multi-threaded BLAS and LAPACK, the difference in
+wall-clock time can be even larger than in CPU time.  The speed of BayesPy for
+these models could potentially be improved by using collapsed or stochastic
+variational inference, and the accuracy could be improved by using
 Gaussian-Wishart nodes which do not factorize with respect to the Gaussian and
@@
 %
-Future plans for BayesPy include implementing more inference engines (e.g.,
-maximum likelihood, expectation propagation and Gibbs sampling), improving the
-VB engine (e.g., collapsed variational inference \citep{Hensman:2012} and
-Riemannian conjugate gradient method \citep{Honkela:2010}) and implementing nodes
-for non-parametric models (e.g., Gaussian processes and Dirichlet processes).
+In addition to the standard variational message passing, it supports several
+advanced methods such as stochastic and collapsed variational inference.  Future
+plans include support for non-conjugate models and non-parametric nodes (e.g.,
+Gaussian and Dirichlet processes).
+% Future plans for BayesPy include implementing more inference engines (e.g.,
+% maximum likelihood, expectation propagation and Gibbs sampling), improving the
+% VB engine (e.g., collapsed variational inference \citep{Hensman:2012} and
+% Riemannian conjugate gradient method \citep{Honkela:2010}) and implementing nodes
+% for non-parametric models (e.g., Gaussian processes and Dirichlet processes).
 

### 2015-03-13 response.tex
Almost ready revised version

@@
 
-I would like to thank..
+I want to express my gratitude to the action editor and all three reviewers for
+their in-depth evaluation and constructive comments.  I have taken into account
+the reviews and revised the paper.  In addition, I have added some new features
+to BayesPy and released a new version (0.3.1).  The four major changes in this
+revised submission can be listed as:
+\begin{itemize}
 
-added features etc
+\item BayesPy now supports several advanced VB techniques: stochastic
+  variational inference, collapsed inference, Riemannian conjugate gradient
+  learning, pattern searches and deterministic annealing.
+
+\item The extremely simple example in the previous submission has been replaced
+  by a Gaussian mixture model example.
+
+\item A detailed comparison to Infer.NET has been added.
+
+\item The reported installation issues have been fixed.
+
+\end{itemize}
+More detailed responses for each reviewer are given below.
 
@@
 
-\item \underline{syntax}\\
-  yeah yeah
+\item \underline{syntax}
+  \\
+  The API in the online documentation (bayespy.org) provides details about the
+  syntax for nodes and methods.  For instance, more details about the syntax of
+  the gating node has been added in the documentation.  However, it is typically
+  better to use Mixture node directly to achieve mixture distributions, as shown
+  in the example in section 3.
 
-\item \underline{speed comparison}\\
-  ooh yeeeeah
+\item \underline{speed comparison}
+  \\
+  The revised version contains a more detailed comparison with Infer.NET.  The
+  other relevant active packages are basically MCMC packages.  Because it is not
+  obvious how to weigh different aspects of MCMC and VB methods in general, a
+  comparison with the MCMC packages is out of the scope of this paper.  However,
+  because Infer.NET seems to offer practically the same functionality, section 4
+  provides a detailed comparison between BayesPy and Infer.NET, including a
+  speed comparison.
 
-\item \underline{supported models}\\
+\item \underline{supported models}
+  \\
+  The revised version makes it more clear that BayesPy supports conjugate
+  exponential family models and does not currently support non-conjugate models.
+  An example of a non-supported model is given in section 4.  Examples of
+  supported popular models are listed in section 2.  Section 3 also provides an
+  example implementation of a Gaussian mixture model, which is much more
+  interesting model than the very simple model in the previous version of the
+  paper.  Because of the page limit, more examples are provided in the online
+  documentation.  Section 4 also discusses Gaussian mixture and PCA models,
+  which are given implementations in the supplementary material.  The results in
+  the section also provide some rough idea about computational costs and how
+  large models can be constructed without using stochastic variational
+  inference.
 
-\item \underline{scheduling}\\
+\item \underline{scheduling}
+  \\
+  The updates (and all computations) are performed sequentially, that is,
+  variables are updated in turns.  This is now mentioned in Section 3.  Parallel
+  computing is not supported.
 
@@
 
-\item \underline{installation instructions}\\
+\item \underline{installation}
+  \\
+  The installation instructions in the online documentation (bayespy.org) have
+  been improved.  Because BayesPy itself is written purely in Python and thus
+  does not need compiling, the only challenge is in installing the requirements
+  (a recent NumPy stack).  In order to make the installation of the NumPy stack
+  much easier, the instructions mention the possibility to use pre-built
+  versions of the NumPy stack (e.g., Anaconda or Enthought).  This is quite
+  close to having a stand-alone executable as suggested.
 
-\item \underline{user community}\\
+  The reviewer was not able to use the package because the execution halted.
+  This was caused by an unfortunate regression bug in matplotlib 1.4.0 and a
+  workaround for it in BayesPy.  These issues have now been fixed.  I want to
+  thank the reviewer for reporting the bug.
+
+  The error in the second installation seems to be caused by a server error in
+  PyPI website (or altenatively some issue in matplotlib), according to my
+  understanding.  I hope the reviewer does not experience this issue anymore.
+
+\item \underline{section headings}
+  \\
+  Section 2 is now labelled ``Features''.
+
+\item \underline{user community}
+  \\
+  It is difficult to give any accurate estimates of the size of the user
+  community.  However, based on given stars, the number of forks and reported
+  issues on Github, there seems to be interested users.  In addition, the number
+  of downloads in pypi.org and mloss.org give some idea about the number of
+  users.
 
 \item \underline{speed comparison}
+  \\
+  The revised version contains a more detailed comparison with Infer.NET.  The
+  other relevant active packages are basically MCMC packages.  It is not obvious
+  how to weigh different aspects of MCMC and VB methods in general, thus a
+  comparison to the MCMC packages is out of the scope of this paper.  However,
+  because Infer.NET seems to offer practically the same functionality, section 4
+  provides a detailed comparison between BayesPy and Infer.NET, including a
+  speed comparison.
 
@@
 
-Reviewer 3 did not point out any BLAH BLAH BLAH
+Reviewer 3 did not point out any modifications the author should make.
 

### 2015-03-13 bayespy.tex
Almost ready revised version

@@
 \begin{keywords}
-  Variational Bayes, variational message passing, Python, probabilistic
-  programming
+  Variational Bayes, probabilistic programming,
+  %variational message passing,
+  Python
 \end{keywords}
@@
 
-BayesPy is a Python 3 package providing tools for constructing models and
-performing VB inference easily and efficiently.  It is based on the variational
-message passing (VMP) framework which defines a simple message passing protocol
-\citep{Winn:2005}.  This enables implementation of small general nodes that can
-be used as building blocks for large complex models.  BayesPy offers a
-comprehensive collection of built-in nodes that can be used to build a wide
-range of models and a simple interface for implementing custom nodes.  The
-package is released under the GNU General Public License v3.0 (GPLv3).
+BayesPy is a Python 3 package providing tools for constructing conjugate
+exponential family models and performing VB inference easily and efficiently.
+It is based on the variational message passing (VMP) framework which defines a
+simple message passing protocol \citep{Winn:2005}.  This enables implementation
+of small general nodes that can be used as building blocks for large complex
+models.  BayesPy offers a comprehensive collection of built-in nodes that can be
+used to build a wide range of models and a simple interface for implementing
+custom nodes.  The package is released under the GNU General Public License v3.0
+(GPLv3).
 
@@
 Nodes are the key part of BayesPy as they are used to construct the model.
-There are two types of nodes in BayesPy: stochastic and deterministic.
-Stochastic nodes correspond to probability distributions and deterministic nodes
-correspond to functions.  Built-in stochastic nodes include all common
-exponential family distributions (e.g., Gaussian, gamma and Dirichlet
-distributions), a general mixture distribution and a few complex nodes for
-dynamic variables (e.g., discrete and Gaussian Markov chains).  Built-in
-deterministic nodes include a gating node and a general sum-product node.
+BayesPy has two types of nodes: stochastic and deterministic.  Stochastic nodes
+correspond to probability distributions and deterministic nodes correspond to
+functions.  Built-in stochastic nodes include all common exponential family
+distributions (e.g., Gaussian, gamma and Dirichlet distributions), a general
+mixture distribution and a few complex nodes for dynamic variables (e.g.,
+discrete and Gaussian Markov chains).  Built-in deterministic nodes include a
+gating node and a general sum-product node.
 %The
@@
 \citep{Katahira:2008}, collapsed inference \citep{Hensman:2012},
-Riemannian-gradient-based optimization \citep{Honkela:2010}, parameter
+Riemannian conjugate gradient learning \citep{Honkela:2010}, parameter
 expansions \citep{Qi:2007} and pattern searches \citep{Honkela:2003}.  For
@@
 release can be installed from Python Package Index (PyPI) and detailed
-instructions can be found from the comprehensive online documentation
-\citep{bayespy.org}.  The latest development version is available at
+instructions can be found from the comprehensive online
+documentation\footnote{\url{http://bayespy.org}}.  The latest development
+version is available at
 GitHub\footnote{\url{https://github.com/bayespy/bayespy}}, which is also the
@@
 
-This section demonstrates BayesPy with a Gaussian mixture model.  A simple
-artificial dataset is created by drawing 200 samples from a 2-dimensional
-Gaussian distribution.  80 samples have mean $[1, 1]$ and 120 samples have mean
-$[0, 0]$:
+This section demonstrates the key steps in using BayesPy.  An artificial
+Gaussian mixture dataset is created by drawing 500 samples from two
+2-dimensional Gaussian distributions.  200 samples have mean $[2, 2]$ and 300
+samples have mean $[0, 0]$:
 \begin{lstlisting}
 import numpy as np
-N = 200, D = 2
+N = 500; D = 2
 data = np.random.randn(N, D)
-data[:80,:] += np.ones(D)
+data[:200,:] += 2*np.ones(D)
 \end{lstlisting}
@@
 mu = nodes.Gaussian(np.zeros(D), 0.01*np.identity(D), plates=(K,))
-Lambda = nodes.Wishart(D, 0.01*np.identity(D), plates=(K,))
+Lambda = nodes.Wishart(D, D*np.identity(D), plates=(K,))
 \end{lstlisting}
@@
 \begin{lstlisting}
-y = nodes.Mixture(Z, nodes.Gaussian, mu, Lambda)
+y = nodes.Mixture(z, nodes.Gaussian, mu, Lambda)
 \end{lstlisting}
@@
 Without the random initialization, the clusters would not be separated.  The VMP
-algorithm is run for 100 iterations or until convergence:
+algorithm updates the variables in turns and is run for 200 iterations or until
+convergence:
 \begin{lstlisting}
-Q.update(repeat=100)
+Q.update(repeat=200)
 \end{lstlisting}
-The results can be examined, for instance, by printing the posterior
-distributions of some variables:
+The results can be examined visually by using \texttt{bayespy.plot} module:
+\begin{lstlisting}
+import bayespy.plot as bpplt
+bpplt.gaussian_mixture_2d(y, alpha=alpha)
+bpplt.pyplot.show()
+\end{lstlisting}
+It is also possible to print the parameters of the approximate posterior
+distributions:
 \begin{lstlisting}
 print(alpha)
-print(mu)
 \end{lstlisting}
-In addition, module \texttt{bayespy.plot} contains several plotting functions to
-examine the results visually.
+The \texttt{bayespy.plot} module contains also other functions for visually
+examining the distributions.
+%In addition, module \texttt{bayespy.plot} contains several plotting functions to
+%examine the results visually.
 
@@
 the other related active projects implement MCMC.  The main advantages of
-Infer.NET over BayesPy are its support for non-conjugate models and other
-inference engines.  On the other hand, BayesPy is open-source software and
-supports several advanced VB methods (e.g., stochastic variational inference and
-collapsed inference).
+Infer.NET over BayesPy are its support for non-conjugate models (e.g., logistic
+regression) and other inference engines.  On the other hand, BayesPy is
+open-source software and supports several advanced VB methods (e.g., stochastic
+variational inference and collapsed inference).
 
@@
 Gaussian mixture model (GMM) and principal component analysis
-(PCA).\footnote{The scripts for running the experiments are available in the
+(PCA).\footnote{The scripts for running the experiments are available as
   supplementary material.}  Both models were run for small and large
@@
 The results are summarized in Table~\ref{tab:speed}.  On the smaller datasets,
-Infer.NET is slightly faster because BayesPy has some overhead of Python.
-However, on the larger datasets, BayesPy is significantly faster because it uses
-BLAS and LAPACK libraries for fast numerical operations.\footnote{Assuming that
-  NumPy uses BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}
-In addition, if using multi-threaded BLAS and LAPACK, the difference in
-wall-clock time can be even larger than in CPU time.  The speed of BayesPy for
-these models could potentially be improved by using collapsed or stochastic
-variational inference, and the accuracy could be improved by using
+Infer.NET is slightly faster because BayesPy has some Python overhead.  However,
+the total CPU time for both methods is only a few seconds.  On the larger
+datasets, BayesPy is significantly faster because it uses BLAS and LAPACK
+libraries for fast numerical operations.\footnote{Assuming that NumPy uses
+  BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}  In
+addition, if using multi-threaded BLAS and LAPACK on a multi-core computer, the
+difference in wall-clock time can be even larger than in CPU time.  The speed of
+BayesPy for these models could potentially be improved by using collapsed or
+stochastic variational inference.  The accuracy could be improved by using
 Gaussian-Wishart nodes which do not factorize with respect to the Gaussian and
@@
 
-% related active software projects discussed in the introduction,
-
-% Because the other
-% active Bayesian software packages discussed in the introduction do not implement
-% the variational message passing algorithm,
-
-%* compiled
-
-%* non-conjugate
-
-%* other engines
-
-%* open source
-
-%* advanced
-
-%* dynamic nodes
-
-% \begin{figure}
-%   \centering
-%   \begin{tikzpicture}
-%     \node [obs] (yn) {$y_n$};
-%     \node [latent, above left=of yn] (mu) {$\mu$};
-%     \node [latent, above right=of yn] (tau) {$\tau$};
-
-%     \edge {mu, tau} {yn};
-
-%     \tikzstyle{plate caption} = [caption, node distance=0, inner sep=0pt,
-%     below=5pt of #1.south]
-
-%     \plate {} {(yn)} {$n=1,\ldots,10$};
-%   \end{tikzpicture}
-%   \caption{Graphical model of the example problem.}
-%   \label{fig:graphical_model}
-% \end{figure}
-% In order to demonstrate BayesPy, this section solves an extremely simple problem
-% but which includes the main steps of using BayesPy.  The task is to estimate the
-% unknown mean and precision parameters of a Gaussian distribution given ten
-% observations.  Thus, the likelihood function is Gaussian:
-% \begin{align*}
-%   p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
-% \end{align*}
-% where the Gaussian distribution is parameterized by precision instead of
-% variance.  The unknown mean $\mu$ and precision $\tau$ are given broad Gaussian
-% and gamma priors, respectively:
-% \begin{align*}
-%   p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}), & p(\tau) &= \mathcal{G}(\tau|10^{-6},
-%   10^{-6}).
-% \end{align*}
-% Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
-% The model is constructed in BayesPy by creating nodes:
-% \begin{lstlisting}
-% from bayespy.nodes import GaussianARD, Gamma
-% mu = GaussianARD(0, 1e-6)
-% tau = Gamma(1e-6, 1e-6)
-% \end{lstlisting}
-% These two nodes represent the two unknown variables that we are interested in.
-% They are the parameters of the Gaussian distribution of the observed variable:
-% \begin{lstlisting}
-% y = GaussianARD(mu, tau, plates=(10,))
-% \end{lstlisting}
-% The number of observations is given as plates which define the number of
-% repetitions.  Note that the syntax of the model construction is similar to the
-% mathematical formulation which defines the conditional distributions for each
-% variable.  Thus, it is easy to understand the model from the Python code.  After
-% the model has been constructed, some nodes are marked as observed and given
-% data:
-% \begin{lstlisting}
-% y.observe([4.5, 3.9, 6.3, 5.6, 4.9, 2.8, 7.4, 6.1, 4.8, 2.1])
-% \end{lstlisting}
-% Next, we want to find the posterior approximation for our latent variables.  We
-% create the VB inference engine:
-% \begin{lstlisting}
-% from bayespy.inference import VB
-% Q = VB(y, mu, tau)
-% \end{lstlisting}
-% The inference engine takes all the stochastic nodes of the model as input.  The
-% message passing algorithm can be run by calling \texttt{update} method:
-% \begin{lstlisting}
-% Q.update(repeat=100)
-% \end{lstlisting}
-% The algorithm is run for 100 iterations or until it converges.  In this case,
-% the algorithm converges in four iterations.  The posterior approximation can be
-% examined, for instance, by plotting the probability density functions of the
-% latent variables:
-% \begin{lstlisting}
-% import bayespy.plot as bpplt
-% import numpy as np
-% bpplt.pyplot.figure()
-% bpplt.pdf(Q[mu], np.linspace(0, 10, num=300))
-% bpplt.pyplot.figure()
-% bpplt.pdf(Q[tau], np.linspace(0, 2, num=300))
-% bpplt.pyplot.show()
-% \end{lstlisting}
-% Figure~\ref{fig:posterior} shows the resulting approximate posterior
-% distributions.  The module \texttt{bayespy.plot} contains functions for simple
-% plotting tasks, such as plotting the probability density function or the Hinton
-% diagram of a variable.  Note that the standard \texttt{pyplot} module of
-% Matplotlib is available at the \texttt{plot} module of BayesPy.
-% \begin{figure}
-%   \begin{center}
-%     \begin{subfigure}[b]{0.4\textwidth}
-%       \includegraphics[width=\textwidth]{fig_mu}
-%       \caption{$q(\mu)$}
-%     \end{subfigure}
-%     \begin{subfigure}[b]{0.4\textwidth}
-%       \includegraphics[width=\textwidth]{fig_tau}
-%       \caption{$q(\tau)$}
-%     \end{subfigure}
-%   \end{center}
-%   \caption{The posterior approximations of $\mu$ and $\tau$.}
-%   \label{fig:posterior}
-% \end{figure}
-
 \section{Conclusions}
@@
 advanced methods such as stochastic and collapsed variational inference.  Future
-plans include support for non-conjugate models and non-parametric nodes (e.g.,
+plans include support for non-conjugate models and non-parametric models (e.g.,
 Gaussian and Dirichlet processes).
@@
 \vskip 0.2in
+\small
 \bibliography{bibliography/bibliography.bib}

### 2015-03-13 experiments/make-supplementary.sh

@@ -1,0 +1,4 @@
+#!/bin/bash
+
+tar -czf bayespy-supplementary.tar.gz run_experiments.sh mog_infernet.cs mog_bayespy.py pca_infernet.cs pca_bayespy.py Makefile utils.py
+

### 2015-03-13 experiments/mog_bayespy.py

@@ -1,24 +1,7 @@
 ######################################################################
-# Copyright (C) 2011-2015 Jaakko Luttinen
+# Copyright (C) 2015 Jaakko Luttinen
 #
-# This file is licensed under Version 3.0 of the GNU General Public
-# License. See LICENSE for a text of the license.
-######################################################################
-
-######################################################################
-# This file is part of BayesPy.
-#
-# BayesPy is free software: you can redistribute it and/or modify it
-# under the terms of the GNU General Public License version 3 as
-# published by the Free Software Foundation.
-#
-# BayesPy is distributed in the hope that it will be useful, but
-# WITHOUT ANY WARRANTY; without even the implied warranty of
-# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
-# General Public License for more details.
-#
-# You should have received a copy of the GNU General Public License
-# along with BayesPy.  If not, see <http://www.gnu.org/licenses/>.
+# MIT License
 ######################################################################
 
 import os
@@ -85,9 +68,6 @@ def generate_data(N, D, K, seed=1, spread=3):
 
     np.savetxt('mog-data-%02d.csv' % seed, y, delimiter=',', fmt='%f')
 
-    #plt.plot(y[:,0], y[:,1], 'rx')
-    #plt.show()
-
     return y
 
 
@@ -113,9 +93,6 @@ def run(N=2000, D=3, K=20, seed=1, maxiter=200, spread=3.0):
     # Run inference
     print("Running inference...")
     Q.update(repeat=maxiter, tol=-1e-6)
-
-    print(Q['alpha'])
-    #print(Q['mu'])
 
     v = np.array([Q.L[:(Q.iter+1)], Q.cputime[:(Q.iter+1)]]).T
     np.savetxt("mog-results-%02d-bayespy.csv" % seed, v, delimiter=",")

### 2015-03-13 experiments/mog_infernet.cs

@@ -1,3 +1,6 @@
+// Copyright (c) 2015 Jaakko Luttinen
+// MIT License
+
 using System;
 using System.IO;
 using System.Diagnostics;
@@ -38,9 +41,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                 IfBlock block = Variable.If(evidence);
 
                 // Define a range for the number of mixture components
-                //K = Variable.New<int>().Named("NumComps");
-                //N = Variable.New<int>().Named("NumObs");
-                //D = Variable.New<int>().Named("NumDims");
                 k = new Range(K).Named("k");
                 n = new Range(N).Named("n");
                 d = new Range(D).Named("d");
@@ -119,8 +119,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                 Rand.Restart(seed);
           
                 // Model size
-                //int maxiter = 200;
-                //int K = 40;  // clusters
                 int N = data.Length; // samples
                 int D = data[0].Count; // dimensionality
 
@@ -188,7 +186,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                     }
                     lines.Add(Columns);
                     Row++;
-                    //Console.WriteLine(Row);
                 }
 
                 var array = lines.ToArray();

### 2015-03-13 experiments/pca_bayespy.py

@@ -1,24 +1,7 @@
 ######################################################################
-# Copyright (C) 2011-2015 Jaakko Luttinen
+# Copyright (C) 2015 Jaakko Luttinen
 #
-# This file is licensed under Version 3.0 of the GNU General Public
-# License. See LICENSE for a text of the license.
-######################################################################
-
-######################################################################
-# This file is part of BayesPy.
-#
-# BayesPy is free software: you can redistribute it and/or modify it
-# under the terms of the GNU General Public License version 3 as
-# published by the Free Software Foundation.
-#
-# BayesPy is distributed in the hope that it will be useful, but
-# WITHOUT ANY WARRANTY; without even the implied warranty of
-# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
-# General Public License for more details.
-#
-# You should have received a copy of the GNU General Public License
-# along with BayesPy.  If not, see <http://www.gnu.org/licenses/>.
+# MIT License
 ######################################################################
 
 
@@ -39,42 +22,17 @@ import bayespy.plot as bpplt
 def model(M, N, D):
     # Construct the PCA model with ARD
 
-    if True:
-        # ARD covariance
-        alpha = nodes.Gamma(1e-2,
-                            1e-2,
-                            plates=(D,),
-                            name='alpha')
-        # Loadings
-        W = nodes.GaussianARD(0,
-                              alpha,
-                              shape=(D,),
-                              plates=(M,1),
-                              name='W')
-    elif True:
-
-        # Full covariance
-        alpha = nodes.Wishart(D,
-                              D*np.identity(D),
-                              name='alpha')
-
-        # Loadings
-        W = nodes.Gaussian(np.zeros(D),
-                           alpha,
-                           plates=(M,1),
-                           name='W')
-    else:
-        # Constant covariance
-
-        # Dummy node
-        alpha = nodes.Bernoulli(0.5, name='alpha')
-
-        # Loadings
-        W = nodes.GaussianARD(0,
-                              1,
-                              shape=(D,),
-                              plates=(M,1),
-                              name='W')
+    # ARD covariance
[…]

### 2015-03-13 experiments/pca_infernet.cs

@@ -1,4 +1,7 @@
-﻿using System;
+﻿// Copyright (c) 2015 Jaakko Luttinen
+// MIT License
+
+using System;
 using System.IO;
 using System.Diagnostics;
 using System.Collections.Generic;
@@ -91,7 +94,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                 var ZtimesW = Variable.MatrixMultiply(W, Z).Named("Z*W");
 
                 // Observation noise
-                //pi = Variable.Array<double>(r
                 tau = Variable.GammaFromShapeAndRate(0.01, 0.01);
 
                 // Observations
@@ -124,7 +126,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                 return Distribution<double>.Array(array);
             }
     }
-//*/
 
     public class BayesianPCAModel_Full : BayesianPCAModel
     {
@@ -186,8 +187,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                                                          PositiveDefiniteMatrix.IdentityScaledBy(D, 0.5*D));
                 // Mixing matrix
                 W = Variable.Array<Vector>(rM).Named("W");
-                //W[rM] = Variable.VectorGaussianFromMeanAndPrecision(Vector.Zero(D),
-                //                                                    PositiveDefiniteMatrix.Identity(D)).ForEach(rM);
                 W[rM] = Variable.VectorGaussianFromMeanAndPrecision(Vector.Zero(D),
                                                                     Alpha).ForEach(rM);
 
@@ -289,14 +288,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                 {
                     starttime = get_cputime();
                     bpca.update();
-                    /*
-                    bpca.engine.NumberOfIterations = j+1;
-                    // Infer the marginals
-                    //bpca.engine.Infer(bpca.Alpha);
-                    bpca.engine.Infer(bpca.W);
-                    bpca.engine.Infer(bpca.Z);
-                    bpca.engine.Infer(bpca.tau);
-                    //*/
                     // Measure CPU time
                     endtime = get_cputime();
                     cputime.Add((endtime - starttime) / 1000.0);
@@ -313,9 +304,6 @@ namespace MicrosoftResearch.Infer.Tutorials
                 SaveResults(string.Format("pca-results-{0:00}-infernet.csv", seed),
                             loglike.ToArray(), cputime.ToArray());
                 
[…]

### 2015-03-13 experiments/utils.py

@@ -1,3 +1,10 @@
+######################################################################
+# Copyright (C) 2015 Jaakko Luttinen
+#
+# MIT License
+######################################################################
+
+
 import numpy as np
 import matplotlib.pyplot as plt
 

### 2015-03-13 response.tex
Re-order responses

@@
 
-\item \underline{syntax}
+\item \underline{supported models}
   \\
-  The API in the online documentation (bayespy.org) provides details about the
-  syntax for nodes and methods.  For instance, more details about the syntax of
-  the gating node has been added in the documentation.  However, it is typically
-  better to use Mixture node directly to achieve mixture distributions, as shown
-  in the example in section 3.
+  The revised version makes it more clear that BayesPy supports conjugate
+  exponential family models and does not currently support non-conjugate models.
+  An example of a non-supported model is given in section 4.  Examples of
+  supported popular models are listed in section 2.  Section 3 also provides an
+  example implementation of a Gaussian mixture model, which is much more
+  interesting model than the very simple model in the previous version of the
+  paper.  Because of the page limit, more examples are provided in the online
+  documentation (bayespy.org).  Section 4 also discusses Gaussian mixture and
+  PCA models, which are given implementations in the supplementary material.
+  The results in the section also provide some rough idea about computational
+  costs and how large models can be constructed without using stochastic
+  variational inference.
 
@@
 
-\item \underline{supported models}
-  \\
-  The revised version makes it more clear that BayesPy supports conjugate
-  exponential family models and does not currently support non-conjugate models.
-  An example of a non-supported model is given in section 4.  Examples of
-  supported popular models are listed in section 2.  Section 3 also provides an
-  example implementation of a Gaussian mixture model, which is much more
-  interesting model than the very simple model in the previous version of the
-  paper.  Because of the page limit, more examples are provided in the online
-  documentation.  Section 4 also discusses Gaussian mixture and PCA models,
-  which are given implementations in the supplementary material.  The results in
-  the section also provide some rough idea about computational costs and how
-  large models can be constructed without using stochastic variational
-  inference.
-
 \item \underline{scheduling}
@@
   computing is not supported.
+
+\item \underline{syntax}
+  \\
+  The API in the online documentation provides details about the syntax for
+  nodes and methods.  For instance, more details about the syntax of the gating
+  node has been added in the documentation.  However, it is typically better to
+  use Mixture node directly to achieve mixture distributions, as shown in the
+  example in section 3.
 
@@
 
-\item \underline{section headings}
+\item \underline{speed comparison}
   \\
-  Section 2 is now labelled ``Features''.
+  The revised version contains a more detailed comparison with Infer.NET.  The
+  other relevant active packages are basically MCMC packages.  It is not obvious
+  how to weigh different aspects of MCMC and VB methods in general, thus a
+  comparison to the MCMC packages is out of the scope of this paper.  However,
+  because Infer.NET seems to offer practically the same functionality, section 4
+  provides a detailed comparison between BayesPy and Infer.NET, including a
+  speed comparison.
 
@@
 
-\item \underline{speed comparison}
+\item \underline{section headings}
   \\
-  The revised version contains a more detailed comparison with Infer.NET.  The
-  other relevant active packages are basically MCMC packages.  It is not obvious
-  how to weigh different aspects of MCMC and VB methods in general, thus a
-  comparison to the MCMC packages is out of the scope of this paper.  However,
-  because Infer.NET seems to offer practically the same functionality, section 4
-  provides a detailed comparison between BayesPy and Infer.NET, including a
-  speed comparison.
+  Section 2 is now labelled ``Features''.
 

### 2015-03-16 bayespy.tex
Revised submission

@@
 OpenBUGS \citep{OpenBUGS}, Dimple \citep{Dimple} and Stan \citep{Stan}.  Thus,
-there is a need for an open source and maintained VB software package.
+there is a need for an open-source and maintained VB software package.
 
@@
 Infer.NET over BayesPy are its support for non-conjugate models (e.g., logistic
-regression) and other inference engines.  On the other hand, BayesPy is
-open-source software and supports several advanced VB methods (e.g., stochastic
-variational inference and collapsed inference).
+regression) and other inference engines (EP and Gibbs sampling).  On the other
+hand, BayesPy is open-source software and supports several advanced VB methods
+(e.g., stochastic variational inference and collapsed inference).
 
@@
 difference in wall-clock time can be even larger than in CPU time.  The speed of
-BayesPy for these models could potentially be improved by using collapsed or
-stochastic variational inference.  The accuracy could be improved by using
-Gaussian-Wishart nodes which do not factorize with respect to the Gaussian and
+BayesPy for these models might be improved by using collapsed or stochastic
+variational inference.  The accuracy of BayesPy could be improved by using a
+Gaussian-Wishart node, which does not factorize with respect to the Gaussian and
 Wishart variables.

### 2015-03-16 reply.tex
Revised submission

\documentclass{article}

\usepackage{url}

\title{Review Reply to JMLR Machine Learning Open Source Software Submission (14-324):\\
BayesPy: Variational Bayesian Inference in Python}

\author{Jaakko Luttinen\\
  \url{jaakko.luttinen@aalto.fi}\\
    Aalto University, Finland}

\begin{document}

\maketitle

\section{Responses to the Reviewers of Submission 14-324 (v1)}

I want to express my gratitude to the action editor and all three reviewers for
their in-depth evaluation and constructive comments.  I have taken into account
the reviews and revised the paper.  In addition, I have added some new features
to BayesPy and released a new version (0.3.2).  The four major changes in this
revised submission can be listed as:
\begin{itemize}

\item BayesPy now supports several advanced VB techniques: stochastic
  variational inference, collapsed inference, Riemannian conjugate gradient
  learning, pattern searches and deterministic annealing.

\item The extremely simple example in the previous submission has been replaced
  by a Gaussian mixture model example.

\item A detailed comparison to Infer.NET has been added.

\item The reported installation issues have been fixed.

\end{itemize}
More detailed responses for each reviewer are given below.

% \subsection{Decision (meta-review)}

% The blah blah

% \begin{itemize}

% \end{itemize}

\subsection{Reviewer 1}

\begin{itemize}

\item \underline{supported models}
  \\
  The revised version makes it more clear that BayesPy supports conjugate
  exponential family models and does not currently support non-conjugate models.
  An example of a non-supported model is given in section 4.  Examples of
  supported popular models are listed in section 2.  Section 3 also provides an
  example implementation of a Gaussian mixture model, which is much more
  interesting model than the very simple model in the previous version of the
  paper.  Because of the page limit, more examples are provided in the online
  documentation (bayespy.org).  Section 4 also discusses Gaussian mixture and
  PCA models, for which implementations are given in the supplementary material.
  The results in section 4 also provide some rough idea about computational
  costs and how large models can be constructed (if not using stochastic
  variational inference).

\item \underline{speed comparison}
  \\
  The revised version contains a more detailed comparison with Infer.NET.  The
  other relevant active packages are basically MCMC packages.  Because it is not
  obvious how to weigh different aspects of MCMC and VB methods in general, a
  comparison with the MCMC packages is out of the scope of this paper.  However,
  because Infer.NET seems to offer practically the same functionality, section 4
  provides a detailed comparison between BayesPy and Infer.NET, including a
  speed comparison.

\item \underline{scheduling}
  \\
  The updates (and all computations) are performed sequentially, that is,
  variables are updated in turns.  This is now mentioned in Section 3.  Parallel
  computing is not supported.

\item \underline{syntax}
  \\
  The API in the online documentation provides details about the syntax for
  nodes and methods.  For instance, more details about the syntax of the gating
  node has been added in the documentation.  However, it is typically better to
  use Mixture node directly to achieve mixture distributions, as shown in the
  example in section 3.

\end{itemize}

\subsection{Reviewer 2}

\begin{itemize}

\item \underline{installation}
  \\
  The installation instructions in the online documentation (bayespy.org) have
  been improved.  Because BayesPy itself is written purely in Python and thus
  does not need compiling, the only challenge is in installing the requirements
  (a recent NumPy stack).  In order to make the installation of the NumPy stack
  much easier, the instructions mention the possibility to use pre-built
  versions of the NumPy stack (e.g., Anaconda or Enthought).  This is quite
  close to having a stand-alone executable as suggested.

  The reviewer was not able to use the package because the execution halted.
  This was caused by an unfortunate regression bug in matplotlib 1.4.0 and a
  workaround for it in BayesPy.  These issues have now been fixed.  I want to
  thank the reviewer for reporting the bug.

  The error in the second installation seems to be caused by a server error in
  PyPI website (or altenatively some issue in matplotlib), according to my
  understanding.  I hope the reviewer does not experience this issue anymore.

  Unfortunately, a current bug in Pip or H5py causes some issues to the
  installation but a simple workaround is to install some of the required
  packages manually (see \url{https://github.com/bayespy/bayespy/issues/14}).
  The documentation also suggests installing required packages manually if
  \texttt{pip install bayespy} fails.

\item \underline{speed comparison}
  \\
  The revised version contains a more detailed comparison with Infer.NET.  The
  other relevant active packages are basically MCMC packages.  It is not obvious
  how to weigh different aspects of MCMC and VB methods in general, thus a
  comparison to the MCMC packages is out of the scope of this paper.  However,
  because Infer.NET seems to offer practically the same functionality, section 4
  provides a detailed comparison between BayesPy and Infer.NET, including a
  speed comparison.

\item \underline{user community}
  \\
  It is difficult to give any accurate estimates of the size of the user
  community.  However, based on given stars, the number of forks and reported
  issues on Github, there seems to be interested users.  In addition, the number
  of downloads in pypi.org and mloss.org give some idea about the number of
  users.

\item \underline{section headings}
  \\
  Section 2 is now labelled ``Features''.

\end{itemize}

\subsection{Reviewer 3}

Reviewer 3 did not request any modifications.

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-PDF-mode: t
%%% TeX-master: t
%%% End:

### 2015-05-13 arxiv-submission.sh
Remove tikz and bayesnet dependency

@@ -2,7 +2,7 @@
 
 SOURCE=bayespy
 TARGET=bayespy-arxiv
-SUPP='tikzlibrarybayesnet.code.tex fig_*.pdf jmlr2e.sty'
+SUPP='jmlr2e.sty'
 
 # Remove comments with:
 perl -pe 's/(^|[^\\])%.*/\1%/' < $SOURCE.tex > tmp.tex 

### 2015-06-01 bayespy.tex
Remove tikz and bayesnet dependency
Minor fixes by reviewers

@@
 \usepackage{amsmath}
-\usepackage{tikz}
-\usetikzlibrary{bayesnet}
+%\usepackage{tikz}
+%\usetikzlibrary{bayesnet}
 
@@
 BayesPy can be used to construct conjugate exponential family models.  The
-documentation provides detailed examples of how to construct, for instance,
-principal component analysis models, linear state-space models, mixture models
-and hidden Markov models.  In addition, BayesPy has been used in two
-publications about parameter expansion and time-varying dynamics for linear
-state-space models \citep{Luttinen:2013,Luttinen:2014}.
+documentation provides detailed examples of how to construct a variety of
+models, including principal component analysis models, linear state-space
+models, mixture models and hidden Markov models.  In addition, BayesPy has been
+used in two publications about parameter expansion and time-varying dynamics for
+linear state-space models \citep{Luttinen:2013,Luttinen:2014}.
 % , which show that it is applicable to large complex models requiring efficient
@@
 
-Nodes are the key part of BayesPy as they are used to construct the model.
-BayesPy has two types of nodes: stochastic and deterministic.  Stochastic nodes
-correspond to probability distributions and deterministic nodes correspond to
-functions.  Built-in stochastic nodes include all common exponential family
-distributions (e.g., Gaussian, gamma and Dirichlet distributions), a general
-mixture distribution and a few complex nodes for dynamic variables (e.g.,
-discrete and Gaussian Markov chains).  Built-in deterministic nodes include a
-gating node and a general sum-product node.
+Nodes are the primary building blocks for models in BayesPy.  There are two
+types of nodes: stochastic and deterministic.  Stochastic nodes correspond to
+probability distributions and deterministic nodes correspond to functions.
+Built-in stochastic nodes include all common exponential family distributions
+(e.g., Gaussian, gamma and Dirichlet distributions), a general mixture
+distribution and a few complex nodes for dynamic variables (e.g., discrete and
+Gaussian Markov chains).  Built-in deterministic nodes include a gating node and
+a general sum-product node.
 %The
@@
 
-BayesPy is designed to be simple enough for average users but flexible and
+BayesPy is designed to be simple enough for non-expert users but flexible and
 efficient enough for expert users.  One goal is to keep the syntax easy and
@@
 variational inference \citep{Hoffman:2013}, deterministic annealing
-\citep{Katahira:2008}, collapsed inference \citep{Hensman:2012},
-Riemannian conjugate gradient learning \citep{Honkela:2010}, parameter
-expansions \citep{Qi:2007} and pattern searches \citep{Honkela:2003}.  For
-developers, the unit testing framework helps in finding bugs, making changes and
-implementing new features in a robust manner.
+\citep{Katahira:2008}, collapsed inference \citep{Hensman:2012}, Riemannian
+conjugate gradient learning \citep{Honkela:2010}, parameter expansions
+\citep{Qi:2007} and pattern searches \citep{Honkela:2003}.  For developers, the
+unit testing framework helps in finding bugs, making changes and implementing
+new features in a robust manner.
 
@@
 converged to similar solutions in a similar number of iterations (50--100
-iterations depending on the dataset), we compare the average CPU time per
+iterations depending on the dataset), we compared the average CPU time per
 iteration.

### 2015-06-03 bayespy.tex
Run non-factorized Infer.NET and non-broadcasting BayesPy

@@
 large model used 40-dimensional latent space for 2000 observations with 100
-dimensions.  Infer.NET used a fully factorizing approximation for the PCA model,
-whereas BayesPy did not factorize with respect to the latent space dimensions.
+dimensions.
+For the PCA model, Infer.NET used a fully factorizing approximation and also
+similarly to BayesPy an approximation which did not factorize with respect to
+the latent space dimensions.
+%Infer.NET used a fully factorizing approximation for the PCA model,
+%whereas BayesPy did not factorize with respect to the latent space dimensions.  
 The experiments were run on a Linux computer with a quad-core processor
@@
 
-The results are summarized in Table~\ref{tab:speed}.  On the smaller datasets,
-Infer.NET is slightly faster because BayesPy has some Python overhead.  However,
-the total CPU time for both methods is only a few seconds.  On the larger
-datasets, BayesPy is significantly faster because it uses BLAS and LAPACK
-libraries for fast numerical operations.\footnote{Assuming that NumPy uses
-  BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}  In
-addition, if using multi-threaded BLAS and LAPACK on a multi-core computer, the
-difference in wall-clock time can be even larger than in CPU time.  The speed of
-BayesPy for these models might be improved by using collapsed or stochastic
-variational inference.  The accuracy of BayesPy could be improved by using a
-Gaussian-Wishart node, which does not factorize with respect to the Gaussian and
-Wishart variables.
+The results are summarized in Table~\ref{tab:speed}.  For GMM, BayesPy is
+significantly faster, which is probably mostly explained by the fact that
+BayesPy uses highly optimized numerical libraries (BLAS and LAPACK).  For PCA,
+BayesPy automatically avoids redundant computations which arise because equal
+posterior covariance matrices are shared between latent variables.  This
+broadcasting makes BayesPy significantly faster than Infer.NET.  However,
+broadcasting is not always possible (e.g., if there are missing values in the
+PCA data), thus the table also presents the results when BayesPy is forced not
+to utilize broadcasting.  In that case, BayesPy is a few times slower than
+Infer.NET because Infer.NET uses a fully factorized approximation.  If Infer.NET
+uses similar factorization as BayesPy, Infer.NET may be orders of magnitude
+slower.
+% On the smaller datasets,
+% Infer.NET is slightly faster because BayesPy has some Python overhead.  However,
+% the total CPU time for both methods is only a few seconds.  On the larger
+% datasets, BayesPy is significantly faster because it uses BLAS and LAPACK
+% libraries for fast numerical operations.\footnote{Assuming that NumPy uses
+%   BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}  In
+% addition, if using multi-threaded BLAS and LAPACK on a multi-core computer, the
+% difference in wall-clock time can be even larger than in CPU time.
+The speed of BayesPy for both models might be improved by using collapsed or
+stochastic variational inference.  The accuracy of BayesPy could be improved by
+using a Gaussian-Wishart node, which does not factorize with respect to the
+Gaussian and Wishart variables.
 
@@
   \centering
-  \caption{The average CPU time (in milliseconds) per iteration for each dataset.}
+  \caption{
+    The average CPU time (in milliseconds) per iteration for each
+    dataset.  The results in parentheses have the following meanings: a) BayesPy without utilizing
+    broadcasting. b) Infer.NET using an approximation factorized similarly as in
+    BayesPy.
+  }
   \small
@@
     \hline
-    BayesPy     & 45 & 210  & 52 & 380
+    BayesPy     & 5 & 85  & 7 (74) & 360 (6\,200)
     \\
-    Infer.NET   & 26 & 4400 & 37 & 2200
+    Infer.NET   & 25 & 4\,600 & 37 (350) & 2\,200 (210\,000)
   \end{tabular}

### 2015-06-03 experiments/pca_bayespy.py

@@ -77,7 +77,7 @@ def plot(seed=1, maxiter=None):
     utils.plot('pca', seed, maxiter=maxiter)
 
 
-def run(M=100, N=1000, D=10, seed=42, rotate=False, maxiter=200, debug=False):
+def run(M=100, N=1000, D=10, seed=42, rotate=False, maxiter=200, debug=False, disable_broadcasting=False):
 
     # Generate data
     print("Generating data...")
@@ -87,7 +87,10 @@ def run(M=100, N=1000, D=10, seed=42, rotate=False, maxiter=200, debug=False):
     Q = model(M, N, D)
 
     # Observe data
-    Q['Y'].observe(y)
+    if disable_broadcasting:
+        Q['Y'].observe(y, mask=np.ones((M,N), dtype=np.bool))
+    else:
+        Q['Y'].observe(y)
 
     # Run inference algorithm
     print("Running inference...")
@@ -107,6 +110,7 @@ if __name__ == '__main__':
                                     "d=",
                                     "seed=",
                                     "maxiter=",
+                                    "disable-broadcasting",
                                     "debug",
                                     "rotate"])
     except getopt.GetoptError:
@@ -117,6 +121,7 @@ if __name__ == '__main__':
         print('--rotate         Apply speed-up rotations')
         print('--maxiter=<INT>  Maximum number of VB iterations')
         print('--seed=<INT>     Seed (integer) for the random number generator')
+        print('--disable-broadcasting Do not utilize broadcasting')
         print('--debug          Check that the rotations are implemented correctly')
         sys.exit(2)
 
@@ -128,6 +133,8 @@ if __name__ == '__main__':
             kwargs["maxiter"] = int(arg)
         elif opt == "--debug":
             kwargs["debug"] = True
+        elif opt == "--disable-broadcasting":
+            kwargs["disable_broadcasting"] = True
         elif opt == "--seed":
             kwargs["seed"] = int(arg)
         elif opt in ("--m",):

### 2015-06-03 experiments/pca_infernet.cs

@@ -38,6 +38,7 @@ namespace MicrosoftResearch.Infer.Tutorials
         public VariableArray2D<double> W = null;
         public VariableArray2D<double> Z = null;
         public Variable<double> tau = null;
+        //public VariableArray2D<double> tau = null;
         public VariableArray<double> Alpha = null;
         public Range rN = null;
         public Range rD = null;
@@ -95,10 +96,35 @@ namespace MicrosoftResearch.Infer.Tutorials
 
                 // Observation noise
                 tau = Variable.GammaFromShapeAndRate(0.01, 0.01);
+                //tau = Variable.Array<double>(rM, rN).Named("Tau");
+                //tau[rM,rN] = Variable.GammaFromShapeAndRate(0.01, 0.01).ForEach(rM, rN);
 
                 // Observations
                 data = Variable.Array<double>(rM, rN).Named("Y");
+                /*
+                // Handle missing values
+                bool[,] isMissing = new bool[M,N];
+                for (int i=0; i<M; i++)
+                {
+                    for (int j=0; i<N; i++)
+                    {
+                        isMissing[i,j] = ((i+j) % 2)==0;
+                    }
+                }
+                VariableArray2D<bool> isMissingVar = Variable.Observed<bool>(isMissing, rM, rN);
+                using (Variable.ForEach(rM))
+                {
+                    using (Variable.ForEach(rN))
+                    {
+                        using (Variable.IfNot(isMissingVar[rM,rN]))
+                        {
+                            data[rM, rN] = Variable.GaussianFromMeanAndPrecision(ZtimesW[rM, rN], tau);
+                        }
+                    }
+                }
+                //*/
                 data[rM, rN] = Variable.GaussianFromMeanAndPrecision(ZtimesW[rM, rN], tau);
+                //data[rM, rN] = Variable.GaussianFromMeanAndPrecision(ZtimesW[rM, rN], tau[rM, rN]);
 
                 // End evidence/lowerbound block
                 block.CloseBlock();

### 2015-06-03 experiments/run_experiments.sh

@@ -2,11 +2,13 @@
 
 # PCA experiments
 python pca_bayespy.py --n=500 --m=20 --d=10 --maxiter=200 --seed=1
-# make && mono pca_infernet.exe 10 1 200 0
-make && mono pca_infernet.exe 10 1 200 1
+python pca_bayespy.py --n=500 --m=20 --d=10 --maxiter=200 --seed=1 --disable-broadcasting
+make && mono pca_infernet.exe 10 1 200 1 # full factorization
+make && mono pca_infernet.exe 10 1 200 0 # same factorization as bayespy
 python pca_bayespy.py --n=2000 --m=100 --d=40 --maxiter=200 --seed=2
-# make && mono pca_infernet.exe 40 2 200 0
-make && mono pca_infernet.exe 40 2 200 1
+python pca_bayespy.py --n=2000 --m=100 --d=40 --maxiter=200 --seed=2 --disable-broadcasting
+make && mono pca_infernet.exe 40 2 200 1 # full factorization
+make && mono pca_infernet.exe 40 2 200 0 # same factorization as bayespy
 
 # MoG experiments
 python mog_bayespy.py --n=200 --k=10 --d=2 --maxiter=200 --seed=1 --spread=3.0

### 2015-06-05 bayespy.tex
Revised version

@@
   flexible model construction and efficient inference make BayesPy suitable for
-  both average and expert Bayesian users.
+  both average and expert Bayesian users.  It also supports some advanced
+  methods such as stochastic and collapsed variational inference.
 \end{abstract}
@@
 used to build a wide range of models and a simple interface for implementing
-custom nodes.  The package is released under the GNU General Public License v3.0
-(GPLv3).
+custom nodes.  The package is released under the MIT license.
 
@@
 models, including principal component analysis models, linear state-space
-models, mixture models and hidden Markov models.  In addition, BayesPy has been
-used in two publications about parameter expansion and time-varying dynamics for
-linear state-space models \citep{Luttinen:2013,Luttinen:2014}.
+models, mixture models and hidden Markov models.  BayesPy has also been used in
+two publications about parameter expansion and time-varying dynamics for linear
+state-space models \citep{Luttinen:2013,Luttinen:2014}.
 % , which show that it is applicable to large complex models requiring efficient
@@
 \end{lstlisting}
-The \texttt{plates} keyword argument is used to define repetitions similarly as
+The \texttt{plates} keyword argument is used to define repetitions similarly to
 the plate notation in graphical models.  The cluster assignments are categorical
@@
 (PCA).\footnote{The scripts for running the experiments are available as
-  supplementary material.}  Both models were run for small and large
-artificially generated datasets.  For GMM, the small model used 10 clusters for
-200 observations with 2 dimensions, and the large model used 40 clusters for
-2000 observations with 10 dimensions.  For PCA, the small model used
-10-dimensional latent space for 500 observations with 20 dimensions, and the
-large model used 40-dimensional latent space for 2000 observations with 100
-dimensions.
-For the PCA model, Infer.NET used a fully factorizing approximation and also
-similarly to BayesPy an approximation which did not factorize with respect to
-the latent space dimensions.
+  supplementary material.}  Both models were run for small and large artificial
+datasets.  For GMM, the small model used 10 clusters for 200 observations with 2
+dimensions, and the large model used 40 clusters for 2000 observations with 10
+dimensions.  For PCA, the small model used 10-dimensional latent space for 500
+observations with 20 dimensions, and the large model used 40-dimensional latent
+space for 2000 observations with 100 dimensions.  For the PCA model, Infer.NET
+used both a fully factorizing approximation and also the same approximation as
+in BayesPy which did not factorize with respect to the latent space dimensions.
 %Infer.NET used a fully factorizing approximation for the PCA model,
 %whereas BayesPy did not factorize with respect to the latent space dimensions.  
-The experiments were run on a Linux computer with a quad-core processor
-(i7-4702MQ).  Because the packages run practically identical algorithms and thus
+The experiments were run on a quad-core (i7-4702MQ) Linux computer.
+% with a quad-core processor (i7-4702MQ).
+Because the packages run practically identical algorithms and thus
 converged to similar solutions in a similar number of iterations (50--100
@@
 
-The results are summarized in Table~\ref{tab:speed}.  For GMM, BayesPy is
-significantly faster, which is probably mostly explained by the fact that
-BayesPy uses highly optimized numerical libraries (BLAS and LAPACK).  For PCA,
-BayesPy automatically avoids redundant computations which arise because equal
-posterior covariance matrices are shared between latent variables.  This
-broadcasting makes BayesPy significantly faster than Infer.NET.  However,
-broadcasting is not always possible (e.g., if there are missing values in the
-PCA data), thus the table also presents the results when BayesPy is forced not
-to utilize broadcasting.  In that case, BayesPy is a few times slower than
-Infer.NET because Infer.NET uses a fully factorized approximation.  If Infer.NET
-uses similar factorization as BayesPy, Infer.NET may be orders of magnitude
-slower.
+The results are summarized in Table~\ref{tab:speed}.  For all datasets, BayesPy
+is faster probably because it uses highly optimized numerical libraries (BLAS
+and LAPACK).  For PCA, BayesPy also automatically avoids redundant computations
+which arise because latent variables have equal posterior covariance matrices.
+% This broadcasting makes BayesPy significantly faster than Infer.NET.
+However, such broadcasting is not always possible (e.g., if there are missing
+values in the PCA data).  Thus, the table also presents the results when BayesPy
+is forced to not use broadcasting: BayesPy is slower than Infer.NET on the
+smaller PCA dataset but faster on the larger PCA dataset.
+% Note that Infer.NET uses a fully factorized approximation.
+If Infer.NET used the same factorization for PCA as BayesPy, Infer.NET may be
+orders of magnitude slower.
 % On the smaller datasets,
@@
 % difference in wall-clock time can be even larger than in CPU time.
-The speed of BayesPy for both models might be improved by using collapsed or
-stochastic variational inference.  The accuracy of BayesPy could be improved by
-using a Gaussian-Wishart node, which does not factorize with respect to the
-Gaussian and Wishart variables.
+%
+%The speed of BayesPy for both models might be improved by using collapsed or
+%stochastic variational inference.  The accuracy of BayesPy could be improved by
+%using a Gaussian-Wishart node, which does not factorize with respect to the
+%Gaussian and Wishart variables.
 
@@
   \caption{
-    The average CPU time (in milliseconds) per iteration for each
-    dataset.  The results in parentheses have the following meanings: a) BayesPy without utilizing
-    broadcasting. b) Infer.NET using an approximation factorized similarly as in
+    The average CPU time in milliseconds per iteration for each
+    dataset.  The results in parentheses have the following meanings: a) BayesPy without using
+    broadcasting. b) Infer.NET using the same factorization as in
     BayesPy.
@@
     \hline
-    BayesPy     & 5 & 85  & 7 (74) & 360 (6\,200)
+    BayesPy     & 6 & 90  & 7 (60) & 400 (1\,500)
     \\

### 2015-06-05 reply2.tex
Add reply to reviewers

\documentclass{article}

\usepackage{url}

\title{Review Reply to JMLR Machine Learning Open Source Software Submission (14-324):\\
BayesPy: Variational Bayesian Inference in Python}

\author{Jaakko Luttinen\\
  \url{jaakko.luttinen@aalto.fi}\\
    Aalto University, Finland}

\begin{document}

\maketitle

\section{Responses to the Reviewers of Submission 14-324 (v2)}

Again, I want to express my gratitude to the action editor and the reviewers for
their evaluation and comments.  I have taken into account the reviews and made a
minor revision.  In addition, I have switched to a more permissive license, that
is, from GPLv3 to MIT and released a new version (0.3.3).  Below are my comments
to the issues listed by the action editor:

\begin{itemize}

\item \underline{resolve the different factorisation}

  In the comparison to Infer.NET, I have now also used the same factorization
  for Infer.NET as for BayesPy.  This was not presented previously for a fair
  comparison because Infer.NET was performing extremely badly with the
  factorization used in BayesPy.  Perhaps, Infer.NET is optimized for the full
  factorization.  In any case, the results now summarize both the full
  factorization and the other factorization for Infer.NET.  In addition, because
  of some other updates to BayesPy, the experiments were run again, thus the
  results slightly differ from the previous results.

\item \underline{make the typographic changes}

  I thank for the language suggestions.  I have made the suggested typographic
  changes.

\item \underline{ensure that the install is smooth}

  I have now checked that the example in the paper works with the latest release
  in PyPI and the development version in GitHub.

\end{itemize}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-PDF-mode: t
%%% TeX-master: t
%%% End:

### 2015-06-05 bayespy.tex
Fix minor typos

@@
 used both a fully factorizing approximation and also the same approximation as
-in BayesPy which did not factorize with respect to the latent space dimensions.
+BayesPy which did not factorize with respect to the latent space dimensions.
 %Infer.NET used a fully factorizing approximation for the PCA model,
@@
     dataset.  The results in parentheses have the following meanings: a) BayesPy without using
-    broadcasting. b) Infer.NET using the same factorization as in
+    broadcasting. b) Infer.NET using the same factorization as
     BayesPy.

### 2016-04-04 experiments/pca_infernet.cs
Clean code by removing commented out code

@@ -96,35 +96,10 @@ namespace MicrosoftResearch.Infer.Tutorials
 
                 // Observation noise
                 tau = Variable.GammaFromShapeAndRate(0.01, 0.01);
-                //tau = Variable.Array<double>(rM, rN).Named("Tau");
-                //tau[rM,rN] = Variable.GammaFromShapeAndRate(0.01, 0.01).ForEach(rM, rN);
 
                 // Observations
                 data = Variable.Array<double>(rM, rN).Named("Y");
-                /*
-                // Handle missing values
-                bool[,] isMissing = new bool[M,N];
-                for (int i=0; i<M; i++)
-                {
-                    for (int j=0; i<N; i++)
-                    {
-                        isMissing[i,j] = ((i+j) % 2)==0;
-                    }
-                }
-                VariableArray2D<bool> isMissingVar = Variable.Observed<bool>(isMissing, rM, rN);
-                using (Variable.ForEach(rM))
-                {
-                    using (Variable.ForEach(rN))
-                    {
-                        using (Variable.IfNot(isMissingVar[rM,rN]))
-                        {
-                            data[rM, rN] = Variable.GaussianFromMeanAndPrecision(ZtimesW[rM, rN], tau);
-                        }
-                    }
-                }
-                //*/
                 data[rM, rN] = Variable.GaussianFromMeanAndPrecision(ZtimesW[rM, rN], tau);
-                //data[rM, rN] = Variable.GaussianFromMeanAndPrecision(ZtimesW[rM, rN], tau[rM, rN]);
 
                 // End evidence/lowerbound block
                 block.CloseBlock();
@@ -233,7 +208,7 @@ namespace MicrosoftResearch.Infer.Tutorials
 
                 // End evidence/lowerbound block
                 block.CloseBlock();
-                        
+
                 // Inference engine
                 engine = new InferenceEngine(new VariationalMessagePassing());
 
